% ============================================================================
% Chapter 33 --- Regulatory mapping
% ----------------------------------------------------------------------------
% Purpose : per-provision mapping of the architecture and the playbook to the
%           regulatory and normative perimeter that the international audience
%           of this Reference is expected to encounter. The chapter is the
%           consolidation point of the per-chapter "Regulatory anchoring"
%           pointers added to Ch. 6--14 (layer chapters) and Ch. 18--27 (wave
%           chapters).
% Page target: ~8 pages.
% Dependencies: cited from Ch. 6--14 and Ch. 18--27 via "Regulatory anchoring"
%               paragraphs; complements the catalogue of normative references
%               of Ch. 32.
% ----------------------------------------------------------------------------
% Decisions registered in _coherence-EN.md:
%   D33.1 mapping is per-article / per-control, three columns (provision,
%         architecture element, operationalised in)
%   D33.2 perimeter is international: GDPR + NIS2 + AI Act + Schrems~II +
%         NIST~CSF~2.0 + eIDAS~2 + ISO/IEC~27001/27002/27005 + COBIT~2019
%   D33.3 a national-transposition cautionary note closes the chapter and
%         is referenced from the per-chapter pointers
%   D33.4 mapping is interpretive, not legal advice (caveatbox at chapter top)
% ============================================================================

\chapter{Regulatory mapping}
\label{ch:regulatory-mapping}

\begin{caveatbox}[Status of this chapter]
\noindent This chapter records the working interpretive mapping
between the architecture's contracts (Part~II) and the playbook's
waves (Part~III) on the one hand, and the regulatory provisions
and normative controls most likely to apply to \loc{institution-type}
on the other. The mapping is interpretive and would
need to be reviewed by qualified counsel and by the institution's
data-protection and information-security officers before being
relied on for a compliance argument. National transpositions of
European directives, sectoral regulations, and contractual
obligations from research funders may add provisions that this
chapter does not record.
\end{caveatbox}

The architecture and the playbook of this Reference are not
designed against a regulatory checklist; they are designed against
the engineering question of how a university
can be sovereign at the architectural level, and they are then
read against the regulations that the institution will face. This
chapter performs that second reading at two levels. It opens with the
shared, jurisdiction-neutral \emph{Regulatory Applicability} module,
which abstracts the size- and jurisdiction-invariant
functional-obligation spine (F1--F6) and its
obligation~$\rightarrow$~mechanism matrix
(Table~\ref{tab:reg-matrix}); that matrix is the shared anchor that the
Policy and this Solution both reference for the
Policy~$\leftrightarrow$~Solution traceability. It then continues with
the detailed, per-article EU instantiation that follows the module ---
organised by regulation and, within each regulation, by article or
control. Each per-article mapping row identifies: the provision, the
architecture element (layer, wave, or contract) that the institution
may invoke, and the chapters and sections of this Reference where the
operationalisation is described. The per-article mapping is the
consolidated view of the per-chapter \emph{Regulatory anchoring}
paragraphs introduced in the layer and wave chapters; it is one worked
instantiation --- the EU/EEA regime row of the module's typology
(\S\ref{sec:reg-eu}) --- developed to article granularity.

% --- Shared Regulatory Applicability module (institution-neutral, ------------
% --- jurisdiction-neutral). The SAME module is referenced as a shared annex --
% --- by the Policy (SUITPOLICY). Its obligation->mechanism matrix -------------
% --- (Table~\ref{tab:reg-matrix}) is the shared traceability anchor. ----------
% --- The module uses \section, correctly nesting under this \chapter. ---------
\input{../shared/regulatory-applicability-EN}

\bigskip
\noindent The remainder of this chapter develops the EU/EEA regime row
of the module above (\S\ref{sec:reg-eu}) to per-article granularity,
binding each provision to the specific architecture element and to the
chapter where it is operationalised.

%-----------------------------------------------------------------------------
\section{Method, scope and limitations}
\label{sec:rm-method}

\paragraph{Perimeter.} The regulations and frameworks covered below
are the per-article development of the EU/EEA regime row of the shared
module's typology (\S\ref{sec:reg-eu}); they are those most frequently
encountered by \loc{institution-type} operating
internationally, presented here as one worked instantiation of the
applicable regulatory regime that each adopting institution substitutes
with its own --- the spine and the obligation~$\rightarrow$~mechanism
matrix (\S\ref{sec:reg-matrix}) being identical across regimes, only the
instrument column changing. The worked
instantiation comprises the applicable data-protection regime
(here the EU \gls{gdpr}); the applicable cybersecurity baseline
regime (here the NIS2 Directive); the applicable AI-governance
regulation (here the EU AI Act); the cross-border data-transfer
regime (here the \emph{Schrems~II} judgement and the EDPB transfer
guidelines that operationalise it); the NIST Cybersecurity
Framework~2.0; the applicable electronic-identification and
trust-services regime (here eIDAS~2, the renewed framework for
electronic identification and trust services); the ISO/IEC 27001 /
27002 / 27005 management-system and risk-management family; and
COBIT~2019 as a governance framework. The selection privileges
international applicability over national specificity.

\paragraph{What the mapping is and is not.} The mapping shows
where the architecture's contracts and the playbook's waves
\emph{contribute} to the satisfaction of a regulatory provision.
It does not claim that adopting the architecture is sufficient to
demonstrate compliance: most provisions also require organisational
elements (policies, training, designated officers, audit reports)
that are outside this Reference's scope. It also does not claim
that any specific deployment described in the architecture is
\emph{necessary} for compliance: alternative architectures may
satisfy the same provisions through different means. The mapping
is best read as a tool for identifying which architectural elements
the institution can invoke when responding to a regulator's
question.

\paragraph{Reading conventions.} For each regulation, the
provisions are listed in the order the source instrument adopts.
The architecture-element column references chapters of this
Reference using the \texttt{\textbackslash{}trchap} cross-reference
convention. The
operationalisation column points to specific sections; multiple
references appear when several layers contribute jointly.

\paragraph{Scope of binding force.} Among the references covered,
the data-protection regime, the cybersecurity baseline regime and
the AI-governance regulation are binding statutory instruments
within their regime (in the worked instantiation, the \gls{gdpr},
NIS2 and the AI Act); the cross-border data-transfer regime is
binding authority (here \emph{Schrems~II} case law from the CJEU);
the electronic-identification and trust-services regime is a
binding regulation (here eIDAS~2). NIST~CSF~2.0, ISO/IEC~27001 and
COBIT~2019 are non-binding frameworks unless made binding by
contract (a research funder requirement, a procurement clause) or
by sectoral national transposition. The institution must read its
own contracts and national legal framework alongside this chapter.

%-----------------------------------------------------------------------------
\section{General Data Protection Regulation (GDPR)}
\label{sec:rm-gdpr}

The applicable data-protection regime establishes risk-based
security and accountability obligations applicable to any
institution processing personal data; in the worked instantiation
below this regime is the \gls{gdpr}~\cite{gdpr}. The provisions
below are the most frequently invoked in IT architecture
discussions; the mapping is not exhaustive, and an adopting
institution reads its own data-protection regime in place of the
worked instantiation.

\begin{table}[H]
\centering\small
\caption{GDPR mapping --- principal provisions and architectural
contributions.}
\label{tab:rm-gdpr}
\renewcommand{\arraystretch}{1.20}
\begin{tabularx}{\textwidth}{%
  >{\hsize=0.75\hsize\linewidth=\hsize\bfseries\raggedright\arraybackslash}X%
  >{\hsize=1.05\hsize\linewidth=\hsize\raggedright\arraybackslash}X%
  >{\hsize=1.20\hsize\linewidth=\hsize\raggedright\arraybackslash}X%
}
\toprule
Provision & Architectural contribution & Operationalised in \\
\midrule
Art.~5(1)(f) integrity and confidentiality
  & Identity, policy, network, crypto layers as the technical chain enforcing access control and confidentiality
  & \trchap{ch:identity}{}, \trchap{ch:policy}{}, \trchap{ch:network}{}, \trchap{ch:crypto}{} \\
Art.~6(1)(e)/(f) lawful basis
  & Public-task and legitimate-interest bases for the monitoring, audit and identity processing the architecture entails; the per-processing determination is finalised in the DPIA/RoPA, with the sovereignty-grid institutional dimension recording it
  & \trchap{ch:sovereignty-grid}{}, \trchap{ch:policy}{}, \trchap{ch:observability}{} \\
Art.~25 data protection by design and by default
  & TAC contract enforces minimum necessary attribute release; sovereignty grid surfaces design-time choices
  & \trchap{ch:conceptual-frame}{}, \trchap{ch:sovereignty-grid}{}, \trchap{ch:identity}{} \S\ref{sec:id-contract} \\
Art.~30 records of processing
  & Canonical TAC schema and its persistence in the audit pipeline produce the per-activity record from which RoPA entries can be derived
  & \trchap{ch:observability}{}, \trchap{ch:wave2-observability}{}, \trchap{ch:wave7-audit-ir}{} \\
Art.~32 security of processing
  & Identity, policy, network, crypto, observability layers; conformance suite as evidence of effectiveness
  & \trchap{ch:identity}{}--\trchap{ch:observability}{}, \trchap{ch:conformance}{} \\
Art.~33--34 notification of breaches
  & TAC-aware audit pipeline and the wave-7 IR procedure provide detection and timeline reconstruction
  & \trchap{ch:wave7-audit-ir}{} \\
Art.~35 data-protection impact assessment
  & Sovereignty grid and TAC scenario inventory provide structured input to a DPIA
  & \trchap{ch:sovereignty-grid}{}, \trchap{ch:self-assessment}{} \\
Art.~44--49 international transfers
  & Storage layer placement constraints, cryptographic-key custody, sovereignty-grid data dimension
  & \trchap{ch:storage}{}, \trchap{ch:crypto}{}, \trchap{ch:sovereignty-grid}{} \\
\bottomrule
\end{tabularx}
\end{table}

\paragraph{Scope of the contribution.} The architecture provides
the technical substrate for the obligations above. The
organisational substrate (data-protection officer, RoPA
maintenance, supplier contracts, data-subject-rights procedures)
is outside this Reference's scope and would need to be in place
independently.

%-----------------------------------------------------------------------------
\section{NIS2 Directive}
\label{sec:rm-nis2}

The applicable cybersecurity baseline regime applies to the
classes of entity its scoping rules designate as in scope; in the
worked instantiation below this regime is the NIS2
Directive~\cite{nis2}, whose classification basis distinguishes
essential and important entities. Whether a research-intensive
university falls in scope depends on the applicable national
transposition of that regime (research organisations are an
Annex~II covered sector (important entities), but the Art.~6(41)
definition excludes education establishments, so universities
usually enter scope by another route --- public-administration
designation, digital infrastructure, hospital affiliation ---
depending on the national transposition). The provisions below are most
frequently relevant when the institution is in scope.

\begin{table}[H]
\centering\small
\caption{NIS2 mapping --- governance, risk-management measures and
incident reporting.}
\label{tab:rm-nis2}
\renewcommand{\arraystretch}{1.20}
\begin{tabularx}{\textwidth}{%
  >{\hsize=0.75\hsize\linewidth=\hsize\bfseries\raggedright\arraybackslash}X%
  >{\hsize=1.05\hsize\linewidth=\hsize\raggedright\arraybackslash}X%
  >{\hsize=1.20\hsize\linewidth=\hsize\raggedright\arraybackslash}X%
}
\toprule
Provision & Architectural contribution & Operationalised in \\
\midrule
Art.~20 governance --- management-body accountability and training
  & Wave-0 governance discipline; sovereignty-grid review by the management body
  & \trchap{ch:wave0-foundations}{}, \trchap{ch:cross-cutting}{} \\
Art.~21(2)(a) risk-analysis and information-system security policy
  & Sovereignty grid and self-assessment; cross-cutting risk register
  & \trchap{ch:sovereignty-grid}{}, \trchap{ch:self-assessment}{}, \trchap{ch:cross-cutting}{} \\
Art.~21(2)(b) incident handling
  & TAC-aware audit pipeline and the wave-7 IR procedure
  & \trchap{ch:wave7-audit-ir}{} \\
Art.~21(2)(c) business continuity, backup, crisis management
  & Reversibility playbook; storage and orchestration backup contracts
  & \trchap{ch:reversibility}{}, \trchap{ch:storage}{}, \trchap{ch:orchestration}{} \\
Art.~21(2)(d) supply-chain security
  & Lifecycle wave; SLSA / Sigstore supply-chain instantiation; procurement-clause workstream
  & \trchap{ch:wave9-lifecycle}{}, \trchap{ch:wave0-foundations}{} \\
Art.~21(2)(e) acquisition, development, maintenance security
  & Conformance suite; coexistence-pattern library; orchestration GitOps controls
  & \trchap{ch:conformance}{}, \trchap{ch:coexistence-patterns}{}, \trchap{ch:orchestration}{} \\
Art.~21(2)(f) policies to assess effectiveness
  & Conformance-suite execution and reporting; observability and audit dashboards
  & \trchap{ch:conformance}{}, \trchap{ch:observability}{} \\
Art.~21(2)(g) cyber hygiene and training
  & Endpoint posture controls; identity-layer MFA / FIDO2 baseline
  & \trchap{ch:endpoint}{}, \trchap{ch:identity}{} \\
Art.~21(2)(h) cryptography policies and procedures
  & Cryptographic services contract; HSM custody; key-rotation procedure
  & \trchap{ch:crypto}{} \\
Art.~21(2)(i) HR security, access control, asset management
  & Identity layer (joiner/mover/leaver via SCIM); endpoint inventory; asset-tagging in observability
  & \trchap{ch:identity}{}, \trchap{ch:endpoint}{}, \trchap{ch:observability}{} \\
Art.~21(2)(j) MFA, secured communications
  & Identity-layer MFA contract; cryptographic-services certificate distribution
  & \trchap{ch:identity}{}, \trchap{ch:crypto}{} \\
Art.~23 incident reporting (24h / 72h / 1m)
  & TAC-aware audit pipeline and wave-7 IR procedure provide the timeline-reconstruction substrate
  & \trchap{ch:wave7-audit-ir}{} \\
\bottomrule
\end{tabularx}
\end{table}

\paragraph{Cumulative argument.} The cybersecurity baseline
regime's risk-management provisions are designed to be read as an
integrated set, not satisfied control by control. The architecture's contribution
should be presented in the institution's compliance argument as
the technical foundation on which the organisational measures
(policies, drills, third-party assessments) build.

%-----------------------------------------------------------------------------
\section{EU AI Act}
\label{sec:rm-ai-act}

The applicable AI-governance regulation classifies AI systems by
risk level and imposes graduated obligations; in the worked
instantiation below this regulation is the EU AI Act (Regulation
2024/1689). For a university, three
contexts most frequently apply: research, testing and development of AI prior to placing it on the
market or putting it into service (Art.~2(8)), alongside AI
developed for the sole purpose of scientific research and
development (Art.~2(6));
deployment of AI tools that interact with users or generate
content (Art.~50 transparency obligations); and use or hosting of
general-purpose AI models, which carry baseline provider
obligations (Art.~53) and, where classified as posing systemic
risk (Art.~51), additional obligations (Art.~55).

\begin{table}[H]
\centering\small
\caption{EU AI Act mapping --- principal provisions for \loc{institution-type}.}
\label{tab:rm-aiact}
\renewcommand{\arraystretch}{1.20}
\begin{tabularx}{\textwidth}{%
  >{\hsize=0.75\hsize\linewidth=\hsize\bfseries\raggedright\arraybackslash}X%
  >{\hsize=1.05\hsize\linewidth=\hsize\raggedright\arraybackslash}X%
  >{\hsize=1.20\hsize\linewidth=\hsize\raggedright\arraybackslash}X%
}
\toprule
Provision & Architectural contribution & Operationalised in \\
\midrule
Art.~2(6) research exemption
  & Pilot-wave scoping (Researcher-LLM-API scenario); audit pipeline records the research-only nature of the activity
  & \trchap{ch:wave5-pilot}{}, \trchap{ch:wave7-audit-ir}{} \\
Art.~2(8) pre-market R\&D activity exemption
  & Development and testing activity scoped and recorded as pre-market R\&D, distinct from production deployment (real-world testing excluded)
  & \trchap{ch:wave5-pilot}{}, \trchap{ch:wave7-audit-ir}{} \\
Art.~5 prohibited practices
  & Policy-layer authorisation rules; sovereignty-grid institutional dimension flags incompatible procurement
  & \trchap{ch:policy}{}, \trchap{ch:sovereignty-grid}{} \\
Art.~6 high-risk classification
  & TAC scenario inventory; sovereignty-grid risk register
  & \trchap{ch:conceptual-frame}{}, \trchap{ch:cross-cutting}{} \\
Art.~27 fundamental-rights impact assessment
  & Sovereignty-grid institutional dimension; TAC scenario record
  & \trchap{ch:sovereignty-grid}{}, \trchap{ch:self-assessment}{} \\
Art.~50 transparency obligations
  & Policy layer enforces context labelling; observability records AI-interaction events
  & \trchap{ch:policy}{}, \trchap{ch:observability}{} \\
Art.~53 obligations for general-purpose AI providers
  & Lifecycle wave (model documentation and provenance); supply-chain attestations
  & \trchap{ch:wave9-lifecycle}{} \\
\bottomrule
\end{tabularx}
\end{table}

\paragraph{Research-context caveat.} The Art.~2(6) and Art.~2(8) research
exemptions are narrow and would not in themselves shield AI services
that the institution operates as production tools for staff or
students. The pilot wave (Wave~5) is therefore designed to
distinguish research activity from operational deployment, with
distinct audit records and access patterns; the institution's AI
governance body would be expected to maintain this distinction.

%-----------------------------------------------------------------------------
\section{Schrems~II and the cross-border data posture}
\label{sec:rm-schrems2}

The applicable cross-border data-transfer regime governs when
contractual safeguards alone do not satisfy data sovereignty
because the data falls under a third-country surveillance regime;
in the worked instantiation this regime is the CJEU's
\emph{Schrems~II} judgement~\cite{schrems2020} together with the
EDPB transfer guidelines~\cite{edpb-guidelines} that operationalise
its consequences. The architecture's contribution to the
resulting cross-border posture is concentrated in the data
dimension of the sovereignty grid and in the storage and
cryptographic layers.

\begin{table}[H]
\centering\small
\caption{Cross-border data --- elements of the
post-\emph{Schrems~II} posture and their architectural
counterparts.}
\label{tab:rm-schrems2}
\renewcommand{\arraystretch}{1.20}
\begin{tabularx}{\textwidth}{%
  >{\hsize=0.75\hsize\linewidth=\hsize\bfseries\raggedright\arraybackslash}X%
  >{\hsize=1.05\hsize\linewidth=\hsize\raggedright\arraybackslash}X%
  >{\hsize=1.20\hsize\linewidth=\hsize\raggedright\arraybackslash}X%
}
\toprule
Element & Architectural contribution & Operationalised in \\
\midrule
Transfer impact assessment (TIA)
  & Sovereignty-grid data dimension provides the indicator-level structure for the TIA
  & \trchap{ch:sovereignty-grid}{}, \trchap{ch:self-assessment}{} \\
Supplementary technical measures
  & Cryptographic services contract; storage-layer key-custody constraints; per-tenant key separation
  & \trchap{ch:crypto}{}, \trchap{ch:storage}{} \\
Standard contractual clauses (SCC) supplementation
  & Procurement-clause workstream of Wave~0; sovereignty-grid review of supplier disclosure obligations
  & \trchap{ch:wave0-foundations}{}, \trchap{ch:sovereignty-grid}{} \\
Onward-transfer surveillance
  & Audit pipeline records cross-border traffic and identity-federation interactions
  & \trchap{ch:observability}{}, \trchap{ch:wave7-audit-ir}{} \\
\bottomrule
\end{tabularx}
\end{table}

%-----------------------------------------------------------------------------
\section{NIST Cybersecurity Framework~2.0}
\label{sec:rm-nist-csf}

The NIST CSF~2.0~\cite{nist-csf} is non-binding but widely used as
the structuring framework for a cybersecurity programme. The six
functions (Govern, Identify, Protect, Detect, Respond, Recover)
provide a useful checklist against which the architecture and
the playbook can be presented.

\begin{table}[H]
\centering\small
\caption{NIST CSF~2.0 mapping --- the architecture's contribution
by function.}
\label{tab:rm-nist-csf}
\renewcommand{\arraystretch}{1.20}
\begin{tabularx}{\textwidth}{%
  >{\hsize=0.75\hsize\linewidth=\hsize\bfseries\raggedright\arraybackslash}X%
  >{\hsize=1.05\hsize\linewidth=\hsize\raggedright\arraybackslash}X%
  >{\hsize=1.20\hsize\linewidth=\hsize\raggedright\arraybackslash}X%
}
\toprule
Function & Architectural contribution & Operationalised in \\
\midrule
\textsf{GOVERN} (GV)
  & Sovereignty grid; wave-0 governance discipline; cross-cutting risk register
  & \trchap{ch:sovereignty-grid}{}, \trchap{ch:wave0-foundations}{}, \trchap{ch:cross-cutting}{} \\
\textsf{IDENTIFY} (ID)
  & Self-assessment; institutional-precedents catalogue; coexistence-pattern inventory
  & \trchap{ch:self-assessment}{}, \trchap{ch:institutional-precedents}{}, \trchap{ch:coexistence-patterns}{} \\
\textsf{PROTECT} (PR)
  & Identity, policy, network, endpoint, crypto layers
  & \trchap{ch:identity}{}--\trchap{ch:network}{}, \trchap{ch:endpoint}{}, \trchap{ch:crypto}{} \\
\textsf{DETECT} (DE)
  & Observability layer; TAC-aware audit pipeline
  & \trchap{ch:observability}{}, \trchap{ch:wave2-observability}{} \\
\textsf{RESPOND} (RS)
  & Wave-7 audit and IR; coexistence-pattern fallback procedures
  & \trchap{ch:wave7-audit-ir}{}, \trchap{ch:coexistence-patterns}{} \\
\textsf{RECOVER} (RC)
  & Reversibility playbook; storage backup contract; orchestration declarative redeployment
  & \trchap{ch:reversibility}{}, \trchap{ch:storage}{}, \trchap{ch:orchestration}{} \\
\bottomrule
\end{tabularx}
\end{table}

%-----------------------------------------------------------------------------
\section{eIDAS~2 and the trust-services dimension}
\label{sec:rm-eidas2}

The applicable electronic-identification and trust-services regime
is relevant to \loc{institution-type} in three ways; in
the worked instantiation this regime is the renewed EU framework
for electronic identification and trust services (Regulation
2024/1183 amending Regulation 910/2014). The three ways are:
participation in the EU Digital Identity Wallet ecosystem when
issuing or relying on academic credentials; provision of
qualified electronic signatures for institutional acts; and the
trust-services baseline for cross-border identity federation. The
operationalisation is concentrated in the identity and
cryptographic layers.

\begin{table}[H]
\centering\small
\caption{eIDAS~2 mapping --- elements relevant to \loc{institution-type}
and their architectural counterparts.}
\label{tab:rm-eidas2}
\renewcommand{\arraystretch}{1.20}
\begin{tabularx}{\textwidth}{%
  >{\hsize=0.75\hsize\linewidth=\hsize\bfseries\raggedright\arraybackslash}X%
  >{\hsize=1.05\hsize\linewidth=\hsize\raggedright\arraybackslash}X%
  >{\hsize=1.20\hsize\linewidth=\hsize\raggedright\arraybackslash}X%
}
\toprule
Element & Architectural contribution & Operationalised in \\
\midrule
Trust services (qualified signature, seal, time stamp)
  & Cryptographic services layer; HSM custody; institutional CA pattern
  & \trchap{ch:crypto}{} \\
Notified eID schemes
  & Identity-federation pattern (eduGAIN, eduTEAMS); SAML2 / OIDC carriage
  & \trchap{ch:identity}{} \\
EU Digital Identity Wallet (EUDI Wallet) interoperability
  & Identity layer carrying wallet-issued attestations; policy layer authorising on attribute claims
  & \trchap{ch:identity}{}, \trchap{ch:policy}{} \\
Cross-border recognition obligations
  & Audit pipeline records federation interactions; sovereignty-grid institutional dimension addresses cross-border governance
  & \trchap{ch:observability}{}, \trchap{ch:sovereignty-grid}{} \\
\bottomrule
\end{tabularx}
\end{table}

%-----------------------------------------------------------------------------
\section{ISO/IEC 27001 / 27002 / 27005}
\label{sec:rm-iso}

The ISO/IEC 27000 family provides the management-system framing
that many institutions use to structure their information-security
programme. ISO/IEC 27001:2022 provides the management-system
clauses (clauses 4--10) and the Annex~A control catalogue (93
controls organised in four themes); ISO/IEC 27002:2022 provides
implementation guidance for those controls; ISO/IEC 27005:2022
provides the risk-management methodology. The mapping below
focuses on Annex~A controls that are most directly addressed by
the architecture.

\begin{table}[H]
\centering\small
\caption{ISO/IEC 27001:2022 Annex~A --- controls most directly
addressed by the architecture (illustrative subset).}
\label{tab:rm-iso27001}
\renewcommand{\arraystretch}{1.20}
\begin{tabularx}{\textwidth}{%
  >{\hsize=0.75\hsize\linewidth=\hsize\bfseries\raggedright\arraybackslash}X%
  >{\hsize=1.05\hsize\linewidth=\hsize\raggedright\arraybackslash}X%
  >{\hsize=1.20\hsize\linewidth=\hsize\raggedright\arraybackslash}X%
}
\toprule
Annex~A control & Architectural contribution & Operationalised in \\
\midrule
A.5.7 Threat intelligence
  & Observability ingest; integration with the relevant R\&E ISAC / sectoral threat-intel feeds (e.g.\ REN-ISAC)
  & \trchap{ch:observability}{}, \trchap{ch:wave7-audit-ir}{} \\
A.5.23 Information security for cloud services
  & Sovereignty-grid data and institutional dimensions; procurement-clause workstream
  & \trchap{ch:sovereignty-grid}{}, \trchap{ch:wave0-foundations}{} \\
A.5.30 ICT readiness for business continuity
  & Reversibility playbook; orchestration declarative redeployment
  & \trchap{ch:reversibility}{}, \trchap{ch:orchestration}{} \\
A.8.2 Privileged access rights
  & Identity layer joiner/mover/leaver; policy layer step-up authorisation
  & \trchap{ch:identity}{}, \trchap{ch:policy}{} \\
A.8.5 Secure authentication
  & Identity-layer MFA / FIDO2 baseline; OIDC / SAML2 federation
  & \trchap{ch:identity}{} \\
A.8.15 Logging
  & Observability layer; canonical TAC schema
  & \trchap{ch:observability}{} \\
A.8.16 Monitoring activities
  & Observability dashboards; conformance-suite execution
  & \trchap{ch:observability}{}, \trchap{ch:conformance}{} \\
A.8.22 Network segregation
  & Network-layer zone-based segmentation; Cilium / Open vSwitch instantiation
  & \trchap{ch:network}{} \\
A.8.24 Use of cryptography
  & Cryptographic services contract; key-rotation procedure
  & \trchap{ch:crypto}{} \\
A.8.30 Outsourced development
  & Lifecycle wave; SLSA / Sigstore supply-chain pattern
  & \trchap{ch:wave9-lifecycle}{} \\
A.8.32 Change management
  & Orchestration GitOps reconciler; conformance-suite gating
  & \trchap{ch:orchestration}{}, \trchap{ch:conformance}{} \\
\bottomrule
\end{tabularx}
\end{table}

\paragraph{Risk methodology.} ISO/IEC 27005:2022 provides the
risk-identification, analysis and evaluation methodology against
which the wave-aware risk register of \trchap{ch:cross-cutting}{}
should be cross-checked. The sovereignty grid provides the
indicator-level structure that operationalises several of the
27005 inputs (asset valuation, vulnerability assessment of
supplier dependencies, threat-context evaluation).

%-----------------------------------------------------------------------------
\section{COBIT~2019 governance and management objectives}
\label{sec:rm-cobit}

COBIT~2019 organises governance and management of enterprise IT
into 40 governance and management objectives across five domains
(EDM --- Evaluate, Direct and Monitor; APO --- Align, Plan and
Organise; BAI --- Build, Acquire and Implement; DSS --- Deliver,
Service and Support; MEA --- Monitor, Evaluate and Assess). The
mapping below highlights objectives most directly addressed by
the architecture and the playbook; the institution's IT-governance function
would be expected to consolidate a complete mapping in its IT
governance documentation.

\begin{table}[H]
\centering\small
\caption{COBIT~2019 mapping --- governance and management
objectives most directly addressed (illustrative subset).}
\label{tab:rm-cobit}
\renewcommand{\arraystretch}{1.20}
\begin{tabularx}{\textwidth}{%
  >{\hsize=0.75\hsize\linewidth=\hsize\bfseries\raggedright\arraybackslash}X%
  >{\hsize=1.05\hsize\linewidth=\hsize\raggedright\arraybackslash}X%
  >{\hsize=1.20\hsize\linewidth=\hsize\raggedright\arraybackslash}X%
}
\toprule
Objective & Architectural contribution & Operationalised in \\
\midrule
EDM01 Ensure governance framework
  & Wave-0 governance discipline; cross-cutting management body cadence
  & \trchap{ch:wave0-foundations}{}, \trchap{ch:cross-cutting}{} \\
EDM03 Ensure risk optimisation
  & Sovereignty grid; wave-aware risk register
  & \trchap{ch:sovereignty-grid}{}, \trchap{ch:cross-cutting}{} \\
APO12 Managed risk
  & ISO/IEC 27005 alignment; sovereignty grid as indicator-level structure
  & \trchap{ch:cross-cutting}{}, \trchap{ch:sovereignty-grid}{} \\
APO13 Managed security
  & Layered architecture; conformance suite as effectiveness evidence
  & \trchap{ch:identity}{}--\trchap{ch:crypto}{}, \trchap{ch:conformance}{} \\
BAI03 Managed solutions identification and build
  & Reference instantiations and replaceability conditions
  & \trchap{ch:architectural-principles}{}, layer chapters \\
BAI06 Managed IT changes
  & Orchestration GitOps reconciler; coexistence-pattern controlled migrations
  & \trchap{ch:orchestration}{}, \trchap{ch:coexistence-patterns}{} \\
BAI10 Managed configuration
  & Endpoint declarative configuration; orchestration as configuration source of truth
  & \trchap{ch:endpoint}{}, \trchap{ch:orchestration}{} \\
DSS02 Managed service requests and incidents
  & Wave-7 audit and IR procedure
  & \trchap{ch:wave7-audit-ir}{} \\
DSS04 Managed continuity
  & Reversibility playbook; storage backup contract
  & \trchap{ch:reversibility}{}, \trchap{ch:storage}{} \\
DSS05 Managed security services
  & Identity / policy / network / crypto layers as the security-service backbone
  & \trchap{ch:identity}{}, \trchap{ch:policy}{}, \trchap{ch:network}{}, \trchap{ch:crypto}{} \\
MEA02 Managed system of internal control
  & Conformance suite; observability dashboards; audit pipeline
  & \trchap{ch:conformance}{}, \trchap{ch:observability}{}, \trchap{ch:wave7-audit-ir}{} \\
MEA03 Managed compliance with external requirements
  & This chapter (consolidated regulatory mapping); per-chapter pointers
  & \S\ref{sec:rm-method}--\S\ref{sec:rm-national} \\
\bottomrule
\end{tabularx}
\end{table}

%-----------------------------------------------------------------------------
\section{Applicability beyond the EU jurisdiction}
\label{sec:rm-beyond-eu}

The per-article mapping above is built against EU instruments because
they are the institution's own binding context. It should not be read
as confining the architecture to EU jurisdictions. As established in
\trchap{ch:architectural-principles}{} (\S\ref{sec:ap-necessity}), the
functions the architecture discharges are necessary to any institution
offering a digital infrastructure; that necessity is
jurisdiction-invariant just as it is size-invariant. What changes
across jurisdictions is the instrument that names the obligation, not
the architectural element that contributes to satisfying it. The shared
module that opens this chapter makes this explicit: the regime typology
(\S\ref{sec:reg-typology}) gives worked F1--F6 instantiations for the
EU, the US, the UK, Brazil and South Africa, and the mapping method
(\S\ref{sec:reg-mapping}) gives an adopter in an unlisted jurisdiction a
seven-step procedure to derive its own regime row against the same
invariant matrix. The paragraphs below summarise how a non-EU adopter
reads the per-article EU mapping through that lens.

\paragraph{Lead with the jurisdiction-neutral spine.} A non-EU
institution can ground its compliance argument on the international
frameworks already mapped in this chapter --- ISO/IEC~27001
(\S\ref{sec:rm-iso}), the NIST Cybersecurity Framework~2.0
(\S\ref{sec:rm-nist-csf}) and COBIT~2019 (\S\ref{sec:rm-cobit}) ---
which are adopted worldwide and are independent of any single legal
regime. The EU-instrument mappings then read as one worked example of
how a regional regulatory regime maps onto that spine.

\paragraph{Functional substitution.} Where this chapter cites the
GDPR, NIS2 or the AI Act, a non-EU institution substitutes its host
jurisdiction's functional analogue --- a data-protection regime, a
cybersecurity or critical-infrastructure regime, and an emerging
AI-governance regime respectively. Illustrative, non-exhaustive
analogues for data protection include the UK GDPR and Data Protection
Act~2018, Brazil's LGPD, California's CCPA/CPRA and Canada's PIPEDA;
these are \emph{functional analogues, not legal equivalences}, and the
architectural element invoked against the obligation is unchanged
regardless of which instrument names it.

\paragraph{Verification remains local.} As with national
transpositions (\S\ref{sec:rm-national}), the institution's
data-protection officer, chief information security officer and legal
counsel are the authoritative interpreters of the regime that actually
applies. This section identifies the portable structure of the
argument; it does not assert equivalence between any two legal
instruments.

%-----------------------------------------------------------------------------
\section{National-transposition cautionary note}
\label{sec:rm-national}

The mapping above is constructed against EU regulations and
internationally recognised frameworks. National transpositions
and sectoral regulations frequently add provisions that this
chapter does not record:

\begin{itemize}[leftmargin=1.5em,nosep]
  \item \textbf{Cybersecurity baseline transposition.}
    Jurisdictions vary in the sectoral scope of the applicable
    national transposition of the cybersecurity baseline regime
    (whether higher-education institutions fall under the
    classification basis that designates in-scope entities, whether
    research is in scope), in the national competent authority
    designated to supervise it, and in the incident-reporting
    thresholds and channels.
  \item \textbf{Data-protection national supplements.} Many
    jurisdictions have adopted national data-protection laws that
    operate alongside the general data-protection regime, with
    provisions on research exemptions, employee monitoring, and
    identifiers that the architecture must respect.
  \item \textbf{Sectoral regulations.} University hospitals,
    institutions handling government research contracts, and
    institutions operating critical national infrastructure may
    be subject to sectoral regulations not addressed here.
  \item \textbf{Funder requirements.} Research funders
    (institutional, national, EU framework programmes) frequently
    impose security and data-management obligations through grant
    agreements; these are contractual rather than regulatory but
    operate as binding constraints on the institution.
  \item \textbf{Non-EU jurisdictions.} Institutions operating
    branches or research stations outside the EU are subject to
    the host jurisdiction's regulations; the international scope
    of this Reference does not imply a single legal framework. A
    wholly non-EU institution adopting the architecture reads the
    mapping through \S\ref{sec:rm-beyond-eu}.
\end{itemize}

\noindent The institution's data-protection officer, chief
information security officer, and legal counsel would be the
authoritative interpreters of the provisions that apply at the
institution's specific situation. Where adopters cooperate --- for example through an international
university network --- that forum is a natural locus for sharing
transposition notes and harmonising the practical interpretation of
the framework above.

%-----------------------------------------------------------------------------
\section{Reading the mapping in conjunction with the chapters}
\label{sec:rm-using}

The per-chapter \emph{Regulatory anchoring} paragraphs in the
layer chapters (\trchap{ch:identity}{}--\trchap{ch:orchestration}{})
and the wave chapters
(\trchap{ch:wave0-foundations}{}--\trchap{ch:wave9-lifecycle}{})
each record the principal regulatory references the chapter
contributes to, and point back to the section of this chapter
where the contribution is mapped. The two views are
complementary: the per-chapter pointer answers ``which
regulations does this layer or wave touch~?'', and the
consolidated mapping of this chapter answers ``which layers and
waves contribute to this regulatory provision~?''. A compliance
argument typically draws on both.
