% ============================================================================
% SOLUTION-LAYER ARGUMENTATION MODULE (standalone)
% Sustainable University IT — Technical and Operational Objections and Responses
% Defends: ../solution-EN.tex
%          "A Technical Reference for Architecture and Migration"
%
% Scope: TECHNICAL / OPERATIONAL objections only —
%   architecture complexity, open-source coverage gaps (iOS/Android MDM, ...),
%   sizing realism, brownfield migration and coexistence, vendor lock-in vs
%   replaceability, conformance-test rigour, observability/audit load,
%   reversibility.
%
% Method: questionbox / answerbox / clarificationbox environments,
%   concession-then-rebuttal pattern, and the FIVE-LAYER evidence
%   methodology (legal > standards > sector > empirical > institutional)
%   with citations from the shared bibliography. Fully GENERIC.
% ============================================================================
\documentclass[11pt,a4paper,twoside]{book}

% === Encoding and language ===
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage[english]{babel}

% === Page layout ===
\usepackage[top=2.5cm,bottom=1.75cm,left=2.5cm,right=2.5cm]{geometry}
\usepackage{setspace}
\setstretch{1.0}

% === Typography and fonts ===
\usepackage{lmodern}
\usepackage{microtype}

% === Graphics and colours ===
\usepackage{graphicx}
\usepackage{suit-logo}
\usepackage[dvipsnames,table]{xcolor}
\definecolor{Slate}{RGB}{70,70,90}
\definecolor{Crimson}{RGB}{220,20,60}

% === Tables ===
\usepackage{booktabs}
\usepackage{tabularx}
\usepackage{longtable}
\usepackage{multirow}
\usepackage{array}

% === Lists ===
\usepackage{enumitem}

% === Symbols (checkmarks, crosses) ===
\usepackage{pifont}

% === Links ===
\usepackage[colorlinks=true,linkcolor=NavyBlue,citecolor=OliveGreen,
            urlcolor=BrickRed,plainpages=false,pdfpagelabels]{hyperref}
\usepackage{bookmark}
\bookmarksetup{numbered}
\usepackage{xurl}

% === Fix bookmark level: \section is used directly under \part (no \chapter) ===
\makeatletter
\renewcommand{\toclevel@section}{1}
\renewcommand{\toclevel@subsection}{2}
\renewcommand{\toclevel@subsubsection}{3}
\makeatother

% === Bibliography ===
\usepackage[backend=biber,
            style=numeric,
            sorting=none,
            sortcites=true,
            maxbibnames=99]{biblatex}
\addbibresource{../../shared/suit-consolidated.bib}

% === Citation display mode ===
\usepackage{bib-modes}
\setbibmode{linked}

% === natbib compatibility ===
\let\citep\parencite
\let\citet\textcite

% === Headers ===
\usepackage{fancyhdr}
\setlength{\headheight}{15pt}
\pagestyle{fancy}
\fancyhf{}
\fancyhead[LE,RO]{\thepage}
\fancyhead[RE]{\textit{Sustainable University IT --- Technical Objections \& Responses}}
\fancyhead[LO]{\textit{\leftmark}}
\fancyfoot[C]{\suitfootlogo}
\renewcommand{\headrulewidth}{0.4pt}
\renewcommand{\footrulewidth}{0.4pt}

% === Miscellaneous ===
\usepackage{csquotes}
\usepackage{float}
\usepackage{caption}
\usepackage{tcolorbox}
\tcbuselibrary{skins,breakable}

% === Glossary with tooltips ===
\usepackage{gls-modes}
\input{../../shared/glossary-defs}

% === Localization (key/value layer 0: generic defaults) ===
\usepackage{localization}
\input{../../shared/localization-defaults}

% ============================================================================
% CUSTOM COMMANDS (aligned with the rest of the suite)
% ============================================================================
\newcommand{\keyword}[1]{\textbf{#1}}
\newcommand{\institution}[1]{\textsc{#1}}
\newcommand{\framework}[1]{\textsf{#1}}

% Five-layer evidence markers (legal > standards > sector > empirical > institutional)
\newcommand{\evlegal}{\textcolor{BrickRed}{\textbf{[Legal]}}\space}
\newcommand{\evstd}{\textcolor{NavyBlue}{\textbf{[Standards]}}\space}
\newcommand{\evsector}{\textcolor{Plum}{\textbf{[Sector]}}\space}
\newcommand{\evempirical}{\textcolor{OliveGreen}{\textbf{[Empirical]}}\space}
\newcommand{\evinst}{\textcolor{Slate}{\textbf{[Institutional]}}\space}

% Cross-reference to a chapter of the Technical Reference being defended
\newcommand{\trref}[1]{the Technical Reference (\emph{#1})}

% ============================================================================
% CUSTOM TCOLORBOX ENVIRONMENTS (suite question/answer convention)
% ============================================================================
\newtcolorbox{questionbox}[1][]{
    colback=red!3!white,
    colframe=red!60!black,
    fonttitle=\bfseries,
    title={#1},
    breakable,
    sharp corners=south,
    boxrule=0.6pt,
    left=4pt, right=4pt
}

\newtcolorbox{answerbox}[1][]{
    colback=OliveGreen!3!white,
    colframe=OliveGreen!60!black,
    fonttitle=\bfseries,
    title={#1},
    breakable,
    sharp corners=south,
    boxrule=0.6pt,
    left=4pt, right=4pt
}

\newtcolorbox{clarificationbox}[1][]{
    colback=NavyBlue!3!white,
    colframe=NavyBlue!60!black,
    fonttitle=\bfseries,
    title={#1},
    breakable,
    sharp corners=south,
    boxrule=0.6pt,
    left=4pt, right=4pt
}

% ============================================================================
% SECTION NUMBERING: skip chapter level (we use \part > \section directly)
% ============================================================================
\renewcommand{\thesection}{\arabic{section}}
\renewcommand{\thesubsection}{\thesection.\arabic{subsection}}
\renewcommand{\thesubsubsection}{\thesubsection.\arabic{subsubsection}}

% === Prevent vertical stretching ===
\raggedbottom

% === Part title on right page, content starts immediately on next page ===
\makeatletter
\renewcommand{\@endpart}{\vfil\newpage}
\makeatother

% ============================================================================
\begin{document}

% === Title page ===
\hypersetup{pageanchor=false}
\begin{titlepage}
\centering
\vspace*{0.5cm}

{\suittitlelogo}

\vspace{0.8cm}

{\Huge\bfseries Sustainable University IT}

\vspace{0.5cm}
{\LARGE\itshape Technical and Operational\\Objections and Responses}

\vspace{1.2cm}
{\large An anticipatory argumentation module defending\\
the reference architecture, the sovereignty grid,\\
and the brownfield migration playbook}

\vspace{2cm}

\begin{tcolorbox}[
    colback=NavyBlue!8!white,
    colframe=NavyBlue!85!black,
    boxrule=1.2pt,
    arc=3pt,
    left=12pt, right=12pt, top=8pt, bottom=8pt,
    width=0.88\textwidth,
    halign=center
]
{\large\bfseries Companion to\\
``\suitsolutiontitle\ --- \suitsolutionsubtitlemain\\
\suitsolutionsubtitlerest''}
\end{tcolorbox}

\vspace{1cm}
\suitmetablock

\vspace{0.9cm}

\begin{minipage}{0.85\textwidth}
\centering
{\large\bfseries SUIT}\\[3pt]
{\normalsize\itshape International Working Group\\on Sustainable University IT in the AI Era}\\[4pt]
{\small\itshape (in formation)}
\end{minipage}

\vspace{0.9cm}

\begin{minipage}{0.85\textwidth}
\centering\small\itshape
This module addresses good-faith technical and operational objections
to the Technical Reference. It is intended for chief technology officers,
information-security officers, network and platform engineers, procurement
officers, and the governance bodies that mandate them. Every response is
constructed on five independent evidence layers --- legal, standards,
sector, empirical, and institutional --- and remains generic across
institutions and jurisdictions.
\end{minipage}

\vspace{0.6cm}

\begin{minipage}{0.85\textwidth}
\small\noindent\textbf{Note on the process:}\\
\textit{This document is an always living document. It is provided for
constructive feedback and revision in order to improve its quality. The
engineering of this document was partially supported by digital tools, some
of which incorporate Artificial Intelligence features.}\\[3pt]
\textcolor{Crimson}{\textit{This document contains a portion of imprecisions,
missing content, errors or possibly hallucinations. We believe that a
supportive and cooperative approach by the readers will make its content
quality increase over time.}}
\end{minipage}

\vspace{1.5cm}
\end{titlepage}

% === Frontmatter ===
\frontmatter
\hypersetup{pageanchor=true}

\chapter*{How to read this module}
\addcontentsline{toc}{chapter}{How to read this module}

This module anticipates the \keyword{technical and operational}
objections most likely to be raised against \trref{Architecture and
Migration}. It deliberately excludes political, budgetary-in-principle,
and academic-freedom objections, which are treated in the policy-layer
argumentation; here the question is always the same: \emph{does the
proposed engineering hold up under adversarial scrutiny?}

Each objection is stated in a \textcolor{red!60!black}{\textbf{question
box}} in the words a sceptical engineer or architect would actually use.
The response is given in a \textcolor{OliveGreen!60!black}{\textbf{answer
box}} that follows a strict \keyword{concession-then-rebuttal} pattern:
it first concedes the legitimate core of the objection, then rebuts the
overreach. Where a point needs disambiguation rather than refutation, an
\textcolor{NavyBlue!60!black}{\textbf{clarification box}} is used instead.

\paragraph{The five-layer evidence methodology.}
No response rests on assertion. Each rebuttal is built from up to five
\emph{independent} evidence layers, marked inline so the reader can audit
the chain of support:

\begin{description}[leftmargin=2.2cm,style=nextline]
  \item[\evlegal] binding legal and regulatory instruments of the
        adopter's jurisdiction (\institution{GDPR},
        \institution{NIS2} and sectoral law in the worked EU
        instantiation).
  \item[\evstd] open technical standards and normative frameworks
        (\institution{IETF}, \institution{OASIS}, \institution{NIST},
        \institution{ISO}).
  \item[\evsector] higher-education and research-sector practice,
        toolkits, and sector bodies.
  \item[\evempirical] measured outcomes --- breach data, incident
        reports, performance studies, cost figures.
  \item[\evinst] named institutional precedents operating the relevant
        pattern in production.
\end{description}

A response is considered sound when at least \emph{three} of these layers
converge independently --- the same minimum-three-layer rule applied
throughout the suite.

\tableofcontents

% ============================================================================
\mainmatter
% ============================================================================

\part{Technical and Operational Objections}

% ----------------------------------------------------------------------------
\section{Architecture complexity}
\label{sec:arg-complexity}

\begin{questionbox}[O1.1 --- ``The nine-layer architecture is over-engineered for a university.'']
The reference architecture decomposes the estate into nine layers ---
identity, network, compute, storage, observability, cryptography,
orchestration, policy, endpoint --- each with its own interface contract.
A university is not a hyperscaler. This is enterprise-architecture theatre
that no campus IT team can realistically operate; the layering adds
cognitive load and failure modes without adding value.
\end{questionbox}

\begin{answerbox}[Response O1.1]
\textbf{Concession.} The objection is right that \emph{a large number of
moving parts is itself a risk}. Architectural complexity that is not
matched by operational capacity degrades reliability rather than improving
it; this is the central lesson of dependable-systems engineering, where
each added component is an added fault source unless it is contained by a
clear interface \citep{avizienis2004}.

\medskip
\textbf{Rebuttal.} The objection mistakes \emph{decomposition} for
\emph{addition}. The nine layers are not nine new systems; they are nine
\emph{names for functions that already exist} in every university estate ---
identity, network, storage, and so on are present whether or not they are
drawn as layers. The architecture does not add them; it makes their
\keyword{interfaces explicit} so that complexity becomes \emph{bounded and
auditable} instead of tangled.

\begin{itemize}[leftmargin=1.4em,itemsep=2pt]
  \item \evstd Layered decomposition behind stable interfaces is the
        canonical method for managing, not multiplying, complexity in
        secure systems engineering \citep{nist-sp800-160v2,avizienis2004}.
        \institution{Zero Trust} architecture is itself defined as a set of
        separated planes (policy decision, policy enforcement, data) for
        exactly this reason \citep{nist-zt}.
  \item \evsector The sector's own governance frameworks treat
        function-by-function decomposition as the prerequisite for
        manageability, not its enemy \citep{cobit,dehaes2015}.
  \item \evinst Research-intensive institutions already operate this decomposition in
        production --- separated identity, network, and storage planes are
        visible, for example, at \institution{MIT} \citep{mit-ist,mit-kerberos},
        \institution{Stanford} \citep{stanford-uit,stanford-sunet}, and
        \institution{University of Michigan} \citep{umich-its,umich-core-network}.
        None is a hyperscaler; all operate a layered decomposition along the
        same functional planes, whatever local labels they use.
\end{itemize}

\textbf{Bottom line.} The architecture does not impose nine systems on a
campus that had one. It gives names and contracts to the functions a campus
already runs, so the irreducible complexity is \emph{governed} rather than
\emph{hidden}.
\end{answerbox}

\begin{clarificationbox}[Clarification --- complexity is instantiated to scale]
The nine-layer \emph{model} is invariant; its \emph{instantiation} is not.
A teaching-led institution may realise several layers as a single managed
service or a hosted offering, collapsing operational surface while keeping
the interface contracts intact (\trref{Sizing reference points}). The model
is the vocabulary; the deployment is sized to the institution.
\end{clarificationbox}

\begin{questionbox}[O1.2 --- ``Identity and policy become institution-wide single points of failure, with no availability story --- a commercial SLA beats this.'']
Per-session policy decision routes every activation through the
identity and policy planes. If either is down, nobody works. A
managed commercial identity service at least carries a contractual
SLA for that risk; the Reference proposes to self-operate the most
critical planes of the institution.
\end{questionbox}

\begin{answerbox}[Response O1.2]
\textbf{Concession.} The mediation-point risk is real, and it is
\emph{created by the architecture itself}: making activation
explicit concentrates what a flat estate leaves diffuse. Any
deployment that treated the identity provider or the policy
decision point as ordinary single-instance services would deserve
this objection in full.

\medskip
\textbf{Rebuttal.} The Reference engineers the risk rather than
insuring against it (\trref{Architectural principles}, availability
and failure semantics of the mediation planes): each critical plane
runs redundantly with a declared availability target measured
through the observability layer; policy enforcement points cache
decisions for the activation-token lifetime, so a policy-plane
outage blocks \emph{new} activations without invalidating
established sessions; and each zone class declares its fail mode
(fail-closed for administrative and sensitive zones, fail-degraded
with audit where connectivity availability prevails).

\begin{itemize}[leftmargin=1.4em,itemsep=2pt]
  \item \evstd Zero-trust guidance itself requires exactly this
        engineering: NIST SP~800-207 treats PDP/PEP availability
        and decision caching as design obligations of the model
        \citep{nist-zt}.
  \item \evstd The reference components are built for clustered
        operation --- replicated identity providers, policy engines
        evaluating from locally distributed bundles
        \citep{opa-project} --- so redundancy is configuration, not
        custom engineering.
  \item \evinst The failure semantics are testable, not asserted:
        availability targets are measured by the observability
        layer, and the fail-mode declarations are part of each
        zone's published security profile, exercised through the
        conformance suite (\trref{Conformance and replaceability
        tests}).
\end{itemize}

\textbf{Bottom line.} An SLA is a remedy after the fact; declared,
rehearsed failure semantics are a design. The architecture does not
leave the mediation planes' availability to contract --- it
engineers it, measures it, and makes outage behaviour a
per-zone-class policy choice.
\end{answerbox}

% ----------------------------------------------------------------------------
\section{Open-source coverage gaps}
\label{sec:arg-oss-gaps}

\begin{questionbox}[O2.1 --- ``Open-source MDM cannot manage iOS and Android, so the open-foundations claim is false.'']
The whole document rests on a reference instantiation built on open
foundations. But there is no credible open-source mobile-device-management
stack for \institution{iOS} and \institution{Android} at scale. The moment
a university has a real fleet of phones and tablets, it must buy a
proprietary suite. The ``open foundations'' claim collapses at the endpoint.
\end{questionbox}

\begin{answerbox}[Response O2.1]
\textbf{Concession.} This is the strongest technical objection in the
document, and it is conceded \emph{in full on the facts}. There is no mature
open-source MDM that matches a commercial suite for supervised
\institution{iOS} and managed \institution{Android} at fleet scale. Open
tooling for Apple platforms is real and production-grade for enrolment and
software management \citep{micromdm,munki}, and \institution{osquery}
provides open posture telemetry across operating systems
\citep{osquery}; but a turnkey, vendor-supported mobile MDM on fully open
foundations does not exist today. The Technical Reference says exactly this,
in the open, rather than hiding it.

\medskip
\textbf{Rebuttal.} The objection then over-reaches: it treats one bounded
gap as the collapse of the whole. The open-foundations claim is a claim
about \keyword{replaceability through contracts}, not about purism. The
architecture's test is not ``is every component open?'' but ``can every
component be replaced without re-architecting?'' --- and a proprietary MDM
that exports configuration profiles to a documented format and feeds posture
to the policy plane through a standard interface \emph{passes that test}.

\begin{itemize}[leftmargin=1.4em,itemsep=2pt]
  \item \evstd The endpoint contract is defined by open interfaces, not by
        the product behind them: posture and identity flow over
        \institution{SCIM} \citep{rfc-7643-scim-core}, \institution{OIDC}
        \citep{oidc-core-1.0}, and standard telemetry --- so a proprietary
        suite is a conforming \emph{instantiation}, not an exception.
  \item \evsector The sector already procures exactly this way: the
        \institution{HECVAT} toolkit exists precisely because universities
        buy assessed proprietary endpoint products under a common
        contract \citep{hecvat}. Commercial suites such as
        \citep{jamf,intune} are named in the reference as legitimate
        instantiations for the iOS/Android portion of the estate.
  \item \evempirical Endpoint compromise --- not MDM ideology --- is where
        the measured risk lives: ransomware and credential-driven breaches
        dominate education-sector incident data
        \citep{sophos-ransomware-edu2025,verizon-dbir2024}. The architecture
        prioritises \emph{posture and replaceability} over provenance
        precisely because that is what the breach data rewards.
  \item \evinst Mixed open/proprietary endpoint estates are common at
        research-intensive institutions, which run a dedicated Apple-platform suite
        alongside open configuration management and \institution{osquery}
        for Linux and servers (\institution{MIT}, \institution{Stanford},
        \institution{University of Michigan}
        \citep{mit-ist,stanford-uit,umich-safecomputing}).
\end{itemize}

\textbf{Bottom line.} The iOS/Android MDM gap is real and acknowledged. It
does not falsify the open-foundations claim, because that claim is about
\keyword{contract-bounded replaceability}, and a proprietary MDM that honours
the endpoint interface contract is a conforming instantiation, not a
counter-example.
\end{answerbox}

\begin{questionbox}[O2.2 --- ``Open source is itself the supply-chain risk: unmaintained dependencies, no vendor CVE-response SLA, nobody to sue.'']
The open-foundations default trades a contractual security
relationship for a volunteer one. Open components ship unmaintained
transitive dependencies, carry no committed CVE-response time, and
offer no counterparty bearing liability when something breaks.
\end{questionbox}

\begin{answerbox}[Response O2.2]
\textbf{Concession.} The failure mode is real: abandonware exists,
and an open licence confers no right to a fix. A deployment that
consumed anonymous code with no support arrangement, no provenance
control, and no maintenance assessment would deserve the objection
in full.

\medskip
\textbf{Rebuttal.} The architecture does not consume anonymous
code; it procures open components under the same discipline as any
supplier.

\begin{itemize}[leftmargin=1.4em,itemsep=2pt]
  \item \evsector Where a committed CVE-response time is required,
        the institution buys a commercial support subscription for
        the open component --- procured and assessed like any
        vendor relationship \citep{hecvat}; the counterparty
        exists, and what remains open is the exit.
  \item \evstd Supply-chain provenance is a named control, not a
        hope: SLSA provenance for built artefacts
        \citep{slsa-spec}, with the supply-chain-security ownership
        established for NIS2 Art.~21(2)(d) by the policy module's
        Q8.5.
  \item \evinst The sovereignty grid's \emph{Operational} dimension
        scores maintenance reality --- skills, community,
        certification paths --- per layer \emph{before} adoption
        (\trref{The sovereignty grid}); a component nobody can
        sustain scores itself out.
  \item \evstd ``Nobody to sue'' cuts both ways: a contractual
        remedy does not produce a patch, while source access and
        the replaceability conditions (O5.1) let the institution
        patch, fork, or replace --- bounding the blast radius of
        any single project's abandonment.
\end{itemize}

\textbf{Bottom line.} Supply-chain risk attaches to unmanaged
consumption, not to licence type. The architecture pairs open
components with support contracts where SLAs matter, provenance
controls where builds matter, and an operational-sustainability
score before adoption --- with replaceability as the final
backstop.
\end{answerbox}

% ----------------------------------------------------------------------------
\section{Sizing realism}
\label{sec:arg-sizing}

\begin{questionbox}[O3.1 --- ``The sizing reference points are fantasy; a small university cannot run this.'']
The architecture may suit a flagship research university with hundreds of
network engineers, but it is unaffordable and unstaffable for a small
teaching-led institution. Prescribing a nine-layer estate to a college with
a five-person IT team is sizing fantasy.
\end{questionbox}

\begin{answerbox}[Response O3.1]
\textbf{Concession.} Conceded that \emph{a one-size instantiation would be
fantasy}. A small institution genuinely cannot operate the same realisation
as a multi-petabyte research university, and any document that demanded it
would be unserious about operations.

\medskip
\textbf{Rebuttal.} The architecture does not demand a single realisation. It
distinguishes the \keyword{minimal necessary functional set} --- which is
size-invariant --- from its \keyword{instantiation} --- which scales. The
functions an institution must perform (authenticate subjects, segment the
network, classify and protect data, retain auditable logs) do not disappear
at small scale; they are simply realised more compactly, often as managed or
hosted services. This is a guardrail \emph{against} overclaim, not an
instance of it.

\begin{itemize}[leftmargin=1.4em,itemsep=2pt]
  \item \evlegal The size-invariant core is anchored in law ---
        \loc{data-protection-regime} and
        \loc{cybersecurity-baseline-regime}, worked here through the
        EU instruments: \institution{GDPR}
        imposes accountability, security of processing, and breach
        obligations with \emph{no de-minimis threshold} ---
        a five-person institution owes the same duties as a flagship
        \citep{gdpr}. (The one narrow size-based relief, the
        records-of-processing derogation for organisations below
        250~persons under Article~30(5), is itself conditional and does not
        touch the security, accountability, or breach duties at issue here.)
        \institution{NIS2} likewise scales nothing down: once an
        entity is in scope (by sector-and-size, or by a
        size-independent designation route), its duties apply in
        full regardless of headcount \citep{nis2}.
  \item \evstd The minimal set maps to baseline controls that are stated
        independently of scale \citep{nist-csf,iso27001,nist-zt}; the
        controls are mandatory, the deployment topology is sized.
  \item \evsector \institution{EDUCAUSE} governance guidance frames IT
        provision as a function of mission and risk, explicitly supporting
        shared-service and consortial realisations for smaller institutions
        \citep{educause-governance,educause-isg}.
  \item \evinst National research and education networks
        (e.g.\ \institution{GÉANT}, \institution{RENATER}, \institution{RESTENA}
        in Europe) exist precisely so that small institutions consume identity,
        connectivity, and security as shared services rather than building
        them alone \citep{geant,renater,restena,eduroam,edugain}.
\end{itemize}

\textbf{Bottom line.} The minimal functional set is size-invariant because
the \emph{obligations} are; the instantiation scales down to managed and
shared services. The sizing chapter prescribes envelopes, not a single
mandatory plant.
\end{answerbox}

% ----------------------------------------------------------------------------
\section{Brownfield migration and coexistence}
\label{sec:arg-migration}

\begin{questionbox}[O4.1 --- ``This is a greenfield design; no real university can rip-and-replace its estate.'']
The wave-based playbook reads like a clean-sheet build. Real universities
have decades of accreted systems, in-flight contracts, and departments that
will not be coordinated. A migration that assumes you can stop the world and
rebuild is dead on arrival.
\end{questionbox}

\begin{answerbox}[Response O4.1]
\textbf{Concession.} Conceded that \emph{a rip-and-replace migration would
fail}. Estates are heterogeneous, contracts are mid-term, and federated
governance means no central team can mandate a flag day. Any playbook that
ignored this would be unusable.

\medskip
\textbf{Rebuttal.} The objection describes a design the Technical Reference
explicitly rejects. The playbook is \keyword{brownfield by construction}: it
is wave-based precisely so that each step is independently valuable,
reversible, and \emph{coexists} with incumbent systems. A dedicated
coexistence-patterns library and a reversibility playbook exist for exactly
the heterogeneity the objection raises (\trref{Coexistence patterns library};
\trref{Reversibility playbook}).

\begin{itemize}[leftmargin=1.4em,itemsep=2pt]
  \item \evstd Incremental adoption behind stable interfaces is the
        standard path for cyber-resilient transformation
        \citep{nist-sp800-160v2}; \institution{Zero Trust} migration is
        itself specified as incremental and coexistence-friendly, never as
        a flag day \citep{nist-zt}.
  \item \evsector Sector governance treats migration as portfolio change
        managed against mission risk, not as a single cutover
        \citep{cobit,weill-ross,educause-governance}.
  \item \evempirical The waves are ordered to address the most-exploited
        attack patterns first: identity, segmentation, and observability
        lead because credential abuse dominates measured breach data
        \citep{verizon-dbir2024,ibm-cost-breach-2024} and segmentation
        interrupts its in-network progression \citep{nist-sp800-215}.
        The ordering is
        an evidence-informed prioritisation of the measured exposure, so
        early waves buy down the most-exploited risk before any large
        investment.
  \item \evinst Coexistence is the lived reality at research-intensive
        institutions such as \institution{MIT} and
        \institution{University of Michigan}, where federated
        identity overlays a heterogeneous estate \citep{mit-kerberos,umich-its},
        and \institution{Science DMZ} deployments coexist with the general
        campus network rather than replacing it
        \citep{science-dmz-2013,esnet-cases}.
\end{itemize}

\textbf{Bottom line.} The playbook is the opposite of greenfield: each wave
is independently valuable and reversible, and coexistence with incumbents is
a first-class design goal, not an afterthought.
\end{answerbox}

% ----------------------------------------------------------------------------
\section{Vendor lock-in versus replaceability}
\label{sec:arg-lockin}

\begin{questionbox}[O5.1 --- ``Standardising on a stack just swaps one lock-in for another.'']
You criticise proprietary lock-in, then prescribe a specific open stack ---
a particular identity provider, a particular orchestrator, a particular
storage system. That is still lock-in; you have only changed the vendor. The
``replaceability'' claim is rhetorical.
\end{questionbox}

\begin{answerbox}[Response O5.1]
\textbf{Concession.} Conceded that \emph{naming open products does not by
itself defeat lock-in}. An open component can still trap an institution
through bespoke configuration, undocumented data formats, or unique
operational dependencies. Open licensing is necessary but not sufficient for
replaceability; status-quo bias makes any incumbent --- open or not ---
sticky \citep{samuelson-zeckhauser}.

\medskip
\textbf{Rebuttal.} The architecture does not claim replaceability \emph{from}
choosing open products. It claims replaceability \emph{from} four explicit
conditions enforced by a conformance suite: standard interfaces, open
state and configuration formats, documented exit-data export, and no
hidden dependencies. The
named open instantiation is an \emph{example} that satisfies these
conditions; a proprietary component that satisfies the same conditions is
equally acceptable. Lock-in is defeated by the \keyword{contract}, not by the
\keyword{logo}.

\begin{itemize}[leftmargin=1.4em,itemsep=2pt]
  \item \evstd Every layer's interface is pinned to an open standard ---
        \institution{OIDC}/\institution{SAML}/\institution{SCIM} for identity
        \citep{oidc-core-1.0,saml-core-2.0,rfc-7643-scim-core},
        \institution{KMIP}/\institution{PKCS\#11} for key management
        \citep{kmip-oasis,pkcs11-oasis}, \institution{TLS}~1.3 and
        \institution{ACME} on the wire \citep{rfc-8446-tls13,rfc-8555-acme},
        \institution{OpenTelemetry} and \institution{IPFIX} for telemetry
        \citep{opentelemetry-spec,rfc-7011-ipfix}, and
        \institution{OCI} image formats for workloads \citep{oci-image-spec}.
        Replaceability is the property of conforming to these, regardless of
        the product.
  \item \evlegal Replaceability is also a compliance asset: \institution{GDPR}
        portability and the ability to change processors presuppose that data
        can leave a system in a usable form \citep{gdpr}; standard-contractual
        and processor-change mechanisms assume exit is possible
        \citep{eu-sccs2021}.
  \item \evsector \institution{HECVAT} institutionalises exactly this
        vendor-neutral, contract-based assessment for the sector
        \citep{hecvat}.
  \item \evinst Multi-vendor, standards-pinned identity federations
        (\institution{eduGAIN} and \institution{eduroam} globally, with regional
        federations such as \institution{InCommon})
        demonstrate that replaceability through interfaces is an operating
        reality, not a slogan \citep{edugain,incommon,eduroam}.
\end{itemize}

\textbf{Bottom line.} The defence against lock-in is the four replaceability
conditions and the conformance suite that tests them --- not allegiance to
any product, open or proprietary.
\end{answerbox}

% ----------------------------------------------------------------------------
\section{Conformance-test rigour}
\label{sec:arg-conformance}

\begin{questionbox}[O6.1 --- ``The conformance tests are a checklist, not a real guarantee.'']
``Replaceability conditions'' and a ``conformance day'' sound reassuring, but
a checklist is not proof. Without an enforced, repeatable, and independently
verifiable test, conformance is a self-graded exercise that every vendor will
claim to pass.
\end{questionbox}

\begin{answerbox}[Response O6.1]
\textbf{Concession.} Conceded that \emph{an unenforced checklist is
worthless}. A conformance claim that cannot be re-run by a third party, on
representative data, is marketing. Self-attestation without evidence is a
known failure mode in procurement.

\medskip
\textbf{Rebuttal.} The conformance suite is constructed as an
\keyword{executable, evidence-producing test}, not a questionnaire. It
defines categories of test --- interface fidelity, data portability, exit
rehearsal, and decision auditability --- each of which yields an artefact: a
captured export consumed by an alternative, a replay against a documented
schema, a recorded migration on a representative subset
(\trref{Conformance and replaceability tests}). The test is passed by
\emph{producing the artefact}, not by asserting compliance. Where
sound practice makes a nominal artefact unobtainable ---
non-exportable HSM trust-anchor keys being the canonical case ---
the layer contract declares the substitute artefact (a key-rotation
and re-anchoring rehearsal under dual control), so the gate is
passable without weakening key custody.

\begin{itemize}[leftmargin=1.4em,itemsep=2pt]
  \item \evstd Conformance-to-interface as the unit of verification is the
        standard discipline: the open interfaces themselves
        (\institution{SCIM}, \institution{KMIP}, \institution{OCI},
        \institution{OpenTelemetry}) carry normative conformance criteria
        that a test can mechanically check
        \citep{rfc-7643-scim-core,kmip-oasis,oci-image-spec,opentelemetry-spec}.
  \item \evsector The sector already operates standardised, repeatable
        third-party assessment through \institution{HECVAT}, and research
        cybersecurity centres publish reusable assessment methods
        \citep{hecvat,trustedci}.
  \item \evempirical Tests are weighted toward the data: control categories
        track the attack patterns that actually cause breaches --- exfiltration
        paths, credential reuse, and unpatched transfer software ---
        as reported in incident data
        \citep{verizon-dbir2024,moveit-breach}.
  \item \evinst The artefact-based discipline is already routine in the
        sector, if for a different property: \institution{perfSONAR} and
        \institution{Science DMZ} deployments verify end-to-end behaviour
        against measured baselines rather than vendor claims
        \citep{perfsonar,science-dmz-2013,sciencedmz-perf}. This establishes
        that re-runnable, evidence-producing conformance is an operating
        norm on the research network, the same discipline the replaceability
        suite applies to exit and portability.
\end{itemize}

\textbf{Bottom line.} Conformance is defined as the production of a
re-runnable artefact against an open interface, not a self-graded checklist;
the artefact is what a third party verifies.
\end{answerbox}

% ----------------------------------------------------------------------------
\section{Observability and audit load}
\label{sec:arg-observability}

\begin{questionbox}[O7.1 --- ``Mandating observability everywhere creates a surveillance and cost burden.'']
The observability layer asks for logs, traces, and metrics across the estate.
That is a privacy hazard, a storage cost explosion, and an alert-fatigue
generator. The audit load you create may be worse than the threats you
mitigate.
\end{questionbox}

\begin{answerbox}[Response O7.1]
\textbf{Concession.} Conceded that \emph{indiscriminate collection is harmful}.
Collecting everything is a privacy hazard, a storage liability, and a path to
alert fatigue --- and over-collected personal data is itself a compliance and
breach-exposure risk, not a control.

\medskip
\textbf{Rebuttal.} The objection conflates \emph{indiscriminate} collection
with \emph{purpose-bound, classified} observability. The architecture
specifies observability tied to data classification and retention policy:
what is collected, why, for how long, and who may read it are constrained by
design, not maximised (\trref{Observability and audit}). Auditability is a
legal obligation that exists independently of this document; the architecture
discharges it \emph{proportionately} rather than maximally.

\begin{itemize}[leftmargin=1.4em,itemsep=2pt]
  \item \evlegal Security of processing and breach-detection duties under
        \institution{GDPR} \citep{gdpr} and the technical risk-management
        requirements of \institution{NIS2} \citep{nis2} \emph{require}
        adequate logging --- with the Commission's technical
        articulation for digital-sector entities applied here by
        analogy \citep{eu-regulation-2024-2690};
        the same legal frame caps it through data minimisation and purpose
        limitation \citep{gdpr}. Audit is mandated; over-collection is
        prohibited --- proportionality is the lawful design point.
  \item \evstd Detect-and-respond functions are baseline controls
        \citep{nist-csf,nist-sp800-53,iso27001}; structured, standard
        telemetry (\institution{syslog}, \institution{IPFIX},
        \institution{OpenTelemetry}) makes collection efficient and bounded
        rather than sprawling
        \citep{rfc-5424-syslog,rfc-7011-ipfix,opentelemetry-spec}.
  \item \evempirical Breach economics reward exactly this: faster detection
        and containment measurably reduce breach cost --- breaches
        contained faster cost measurably less, the breach lifecycle
        being a major cost driver
        \citep{ibm-cost-breach-2024,verizon-dbir2024}. Proportionate
        observability is the highest-return control, not overhead.
  \item \evinst Research universities run classification-driven observability
        in production through their security operations and incident-response
        capabilities \citep{umich-safecomputing,stanford-uit}, and the
        sector operates shared trust and incident infrastructure
        (e.g.\ \institution{REN-ISAC} in the US, \institution{FIRST} globally) on this basis
        \citep{ren-isac,first}.
\end{itemize}

\textbf{Bottom line.} The mandate is proportionate, classified observability
--- legally required to exist and legally bounded in extent --- not blanket
surveillance; the standard telemetry interfaces keep its cost bounded.
\end{answerbox}

% ----------------------------------------------------------------------------
\section{Reversibility}
\label{sec:arg-reversibility}

\begin{questionbox}[O8.1 --- ``Reversibility is a comforting promise that never survives contact with production.'']
You promise that each wave is reversible. In practice, once identity,
network, and policy changes are live, rolling back is a fantasy --- the
dependencies entangle, the data has moved, and nobody dares revert. The
reversibility playbook is reassurance, not engineering.
\end{questionbox}

\begin{answerbox}[Response O8.1]
\textbf{Concession.} Conceded that \emph{reversibility is not free and is
often claimed without being engineered}. A rollback that was never designed,
rehearsed, or tested will indeed fail under pressure, and entangled
dependencies can make a naive revert impossible.

\medskip
\textbf{Rebuttal.} That is precisely why reversibility is treated as a
\keyword{designed and rehearsed property}, not an aspiration. Each wave
defines its rollback preconditions, its data-preservation guarantees, and a
tested revert path \emph{before} the wave is committed; coexistence with the
incumbent during a wave is what keeps the revert path open
(\trref{Reversibility playbook}; \trref{Coexistence patterns library}).
Reversibility here is a fault-tolerance requirement, engineered like any
other.

\begin{itemize}[leftmargin=1.4em,itemsep=2pt]
  \item \evstd Designed recoverability and graceful rollback are core
        cyber-resilience and dependable-systems requirements, not optional
        extras \citep{nist-sp800-160v2,avizienis2004}; resilient systems are
        specified to fail back to a known-good state.
  \item \evlegal Reversibility is also a continuity and exit obligation:
        processor-change and data-portability provisions presuppose that a
        configuration can be unwound and data recovered in usable form
        \citep{gdpr,eu-sccs2021}.
  \item \evsector Change managed with explicit rollback gates is standard IT
        governance practice \citep{cobit,dehaes2015}.
  \item \evinst Coexistence-based migrations preserve the revert path in
        production: \institution{Science DMZ} runs alongside the general
        network so the institution can fall back, and federated identity
        overlays --- rather than replaces --- incumbent systems during
        transition
        \citep{science-dmz-2013,esnet-cases,mit-kerberos,umich-its}.
\end{itemize}

\textbf{Bottom line.} Reversibility is engineered and rehearsed per wave, and
kept open by coexistence with the incumbent --- it is a fault-tolerance
requirement, not a comforting promise.
\end{answerbox}

% ============================================================================
\part{Quick-Reference Summary}
% ============================================================================

\section{Objection map}
\label{sec:arg-summary}

The table below summarises every objection in this module, its concession,
the dominant evidence layers, and the one-line rebuttal. \evlegal{}Legal,
\evstd{}Standards, \evsector{}Sector, \evempirical{}Empirical, and
\evinst{}Institutional mark the converging evidence layers used in each
response (minimum three per response).

\renewcommand{\arraystretch}{1.35}
\begin{longtable}{@{}p{0.06\textwidth}p{0.20\textwidth}p{0.20\textwidth}p{0.13\textwidth}p{0.29\textwidth}@{}}
\toprule
\textbf{ID} & \textbf{Objection} & \textbf{Conceded core} &
\textbf{Evidence} & \textbf{One-line rebuttal} \\
\midrule
\endfirsthead
\toprule
\textbf{ID} & \textbf{Objection} & \textbf{Conceded core} &
\textbf{Evidence} & \textbf{One-line rebuttal} \\
\midrule
\endhead
\midrule
\multicolumn{5}{r}{\footnotesize\itshape continued on next page}\\
\endfoot
\bottomrule
\endlastfoot

O1.1 & Architecture is over-engineered for a university &
Unmatched complexity is a real risk &
Std, Sector, Inst &
Layers name functions a campus already runs; contracts \emph{bound} complexity, not add it. \\

O1.2 & Identity/policy are single points of failure; an SLA beats this &
Mediation-point risk is real and architecture-created &
Std, Inst &
Redundancy per plane, PEP caching over token lifetime, declared per-zone fail modes --- engineered, measured, rehearsed. \\

O2.1 & Open-source MDM cannot manage iOS/Android &
No mature open mobile MDM at scale (true) &
Legal, Std, Sector, Emp, Inst &
Open-foundations means contract-bounded replaceability; a conforming proprietary MDM is a valid instantiation. \\

O2.2 & Open source is itself the supply-chain risk &
Abandonware is real; a licence confers no fix &
Std, Sector, Inst &
Support subscriptions where SLAs matter, SLSA provenance, Operational-dimension scoring, replaceability as backstop. \\

O3.1 & Sizing is fantasy for small institutions &
One-size instantiation would be unserious &
Legal, Std, Sector, Inst &
Minimal functional set is size-invariant (no-threshold law); instantiation scales to shared/managed services. \\

O4.1 & Greenfield design; no rip-and-replace is possible &
Rip-and-replace would fail &
Std, Sector, Emp, Inst &
Playbook is brownfield by construction: waves coexist with incumbents and are independently valuable. \\

O5.1 & Open stack just swaps one lock-in for another &
Open licensing alone does not defeat lock-in &
Legal, Std, Sector, Inst &
Replaceability comes from four conditions + conformance tests, not from the logo. \\

O6.1 & Conformance tests are a checklist, not a guarantee &
An unenforced checklist is worthless &
Std, Sector, Emp, Inst &
Conformance = a re-runnable artefact against an open interface, third-party verifiable. \\

O7.1 & Observability is surveillance and cost burden &
Indiscriminate collection is harmful &
Legal, Std, Emp, Inst &
Mandate is proportionate, classified observability --- legally required and legally bounded. \\

O8.1 & Reversibility never survives production &
Un-rehearsed rollback fails &
Legal, Std, Sector, Inst &
Reversibility is designed and rehearsed per wave, kept open by coexistence. \\

\end{longtable}

\bigskip
\noindent\textbf{Methodological note.} Every response in this module concedes
the legitimate core of its objection before rebutting the overreach, and
each rests on at least three independent evidence layers. The recurring
defensive structure --- \emph{size-invariant minimal functional set},
\emph{contract-bounded replaceability}, and \emph{brownfield coexistence} ---
is what allows the architecture to be simultaneously rigorous and applicable
across institutions of very different scale and jurisdiction.

% ============================================================================
\backmatter
% ============================================================================

\printbibliography[heading=bibintoc]

\end{document}
