%============================================================================
% OPERATIONAL EXECUTIVE SUMMARY of the Technical Reference
% Sustainable University IT --- A Technical Reference for Architecture and Migration
%
% Audience: CISO, CTO, infrastructure / network team leads (operational).
% Derived from: ../solution-EN.tex and its chapters (the "Technical Reference").
% Methodology: /companion, applied with an OPERATIONAL framing.
% Fully generic: no institution named.
%============================================================================
\documentclass[11pt,a4paper,twoside]{article}

% --- Encoding and language ---
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage[english]{babel}

% --- Page layout (compact for 3-6 pages) ---
\usepackage[top=1.8cm,bottom=1.6cm,left=2.1cm,right=2.1cm]{geometry}
\usepackage{setspace}
\setstretch{1.04}

% --- Typography ---
\usepackage{lmodern}
\usepackage{microtype}

% --- Graphics and colours ---
\usepackage{graphicx}
\usepackage{suit-logo}
\usepackage[dvipsnames,table]{xcolor}
\definecolor{Slate}{RGB}{70,70,90}
\definecolor{Crimson}{RGB}{220,20,60}
\usepackage{tikz}
\usetikzlibrary{shapes,arrows.meta,positioning,fit,calc,backgrounds}

% --- Tables ---
\usepackage{booktabs}
\usepackage{tabularx}
\usepackage{array}
\usepackage{multirow}

% --- Lists ---
\usepackage{enumitem}
\usepackage{pifont}
\setlist{nosep,leftmargin=1.5em}

% --- Links ---
\usepackage[colorlinks=true,linkcolor=NavyBlue,citecolor=OliveGreen,
            urlcolor=BrickRed]{hyperref}
\usepackage{xurl}

% --- Quotation handling (required by biblatex) ---
\usepackage{csquotes}

% --- Bibliography: biblatex + biber (shared consolidated bib) ---
\usepackage[backend=biber,
            style=numeric,
            sorting=none,
            sortcites=true,
            maxbibnames=99]{biblatex}
\addbibresource{../../shared/suit-consolidated.bib}

% --- Bibliography display modes (standard / annotated / full) ---
\usepackage{bib-modes}
\setbibmode{linked}

% --- Glossary with tooltips ---
\usepackage{gls-modes}
\input{../../shared/glossary-defs}

% === Localization (key/value layer 0: generic defaults) ===
\usepackage{localization}
\input{../../shared/localization-defaults}

% --- Boxes ---
\usepackage{tcolorbox}
\tcbuselibrary{skins,breakable}

% --- Headers ---
\usepackage{fancyhdr}
\setlength{\headheight}{15pt}
\pagestyle{fancy}
\fancyhf{}
\fancyhead[LO,RE]{\small\textit{Operational Summary --- Sustainable University IT}}
\fancyhead[LE,RO]{\small\thepage}
\fancyfoot[C]{\suitfootlogo}
\renewcommand{\headrulewidth}{0.4pt}
\renewcommand{\footrulewidth}{0.4pt}

% --- Semantic commands (aligned with the rest of the suite) ---
\newcommand{\keyword}[1]{\textbf{#1}}
\newcommand{\institution}[1]{\textsc{#1}}
\newcommand{\framework}[1]{\textsf{#1}}

% --- No section numbering ---
\setcounter{secnumdepth}{0}

% --- Compact sections ---
\usepackage{titlesec}
\titlespacing*{\section}{0pt}{0.8em}{0.3em}
\titlespacing*{\subsection}{0pt}{0.5em}{0.2em}
\titleformat{\section}{\normalsize\bfseries\color{NavyBlue}}{}{}{}
\titleformat{\subsection}{\small\bfseries\color{MidnightBlue}}{}{}{}

%----------------------------------------------------------------------------
% Cross-reference helpers back to the full Technical Reference (SUITSOLUTION).
% Self-contained: the summary names the target chapter explicitly so that the
% pointer resolves without building the full Reference first.
%----------------------------------------------------------------------------
% \trchap{N}{title} -> "Chapter N (title) of the Technical Reference"
\newcommand{\trchap}[2]{%
  Chapter~#1%
  \ifx\\#2\\\else\space(\emph{#2})\fi\space of the Technical Reference%
}
% \mainref{...} -> grey italic pointer line into the full Reference
\newcommand{\mainref}[1]{%
  \par\vspace{1pt}
  {\footnotesize\color{gray}\itshape
  $\triangleright$~Full treatment, contracts and references: #1,
  in the Technical Reference~\cite{ng-tech-ref-2026}.}\par\vspace{2pt}}

% --- Operational box palette (consistent with the suite) ---
% principle (NavyBlue) / alert (BrickRed) / action (OliveGreen)
\newtcolorbox{principlebox}[1][]{%
  colback=NavyBlue!5!white, colframe=NavyBlue!70!black,
  fonttitle=\bfseries, title={#1}, boxrule=0.5pt, sharp corners=south,
  left=6pt, right=6pt, top=4pt, bottom=4pt}

\newtcolorbox{alertbox}[1][]{%
  colback=BrickRed!5!white, colframe=BrickRed!70!black,
  fonttitle=\bfseries, title={#1}, boxrule=0.5pt, sharp corners=south,
  left=6pt, right=6pt, top=4pt, bottom=4pt}

\newtcolorbox{actionbox}[1][]{%
  colback=OliveGreen!5!white, colframe=OliveGreen!70!black,
  fonttitle=\bfseries, title={#1}, boxrule=0.5pt, sharp corners=south,
  left=6pt, right=6pt, top=4pt, bottom=4pt}

% --- Prevent vertical stretching ---
\raggedbottom

%============================================================================
\begin{document}

% --- Title block (no separate title page --- space is precious) ---
\begin{center}
\vspace*{-0.2cm}

{\suittitlelogo[0.30\textwidth]}

\smallskip
{\Large\bfseries Sustainable University IT --- Operational Summary}

\smallskip
{\normalsize\itshape Reference Architecture, Migration Waves and Acceptance Gates,
at a Glance}

\smallskip
{\small For the CISO, the CTO, and infrastructure / network team leads}

\smallskip
{\scriptsize Operational companion to: \emph{``Sustainable University IT ---
A Technical Reference for Architecture and Migration''}}

\smallskip
\suitmetablock

\medskip

\begin{tcolorbox}[
    colback=NavyBlue!8!white, colframe=NavyBlue!85!black,
    boxrule=1.2pt, arc=3pt,
    left=10pt, right=10pt, top=8pt, bottom=8pt, halign=center]
{\normalsize\bfseries Nine layers, three planes, ten waves ---
each wave reversible, each closure proven\\[3pt]
by conformance, every component replaceable through a published interface.}
\end{tcolorbox}
\end{center}

\smallskip
\noindent\rule{\textwidth}{0.6pt}

%============================================================================
\section{Operational synopsis}
%============================================================================

\noindent We are standing up a \keyword{vendor-neutral reference
architecture} for university IT and a \keyword{brownfield migration
playbook} to reach it from whatever estate exists today. In operational
terms: a small, strongly protected core (identity, observability, network
backbone) is established first; authorisation is lifted out of scattered
application logic and identity-provider-native rules into a single
\gls{pdp}; the network is carved into declared zones; and every workload,
session and access decision is bound to a \gls{tac} --- a validated bundle of
subject, action, resource and context that the platform can activate,
audit and tear down as a unit. The architecture's normative content is a
set of \keyword{interface contracts} (OIDC/SCIM, RADIUS, NETCONF/YANG, the
Kubernetes and S3 APIs, OpenTelemetry, KMIP/ACME), not a product list:
open-source instantiations prove the contracts are realisable, and any
compliant component --- proprietary or open --- is admissible.
The migration is delivered as \keyword{ten waves} under a dependency graph,
each with a coexistence (shadow / dual-stack) pattern, a declared
reversibility window, and conformance-based exit gates. The pay-off for
operations is concrete: incidents are bounded by TAC instead of cascading
through a flat perimeter, suppliers are swapped through a rehearsed exit
rather than a forklift, and ``done'' is a passing test rather than a
declaration.

\mainref{\trchap{1}{Synthesis}}

%============================================================================
\section{Reference architecture at a glance: nine layers, three planes}
%============================================================================

\begin{table}[h!]
\centering\small
\renewcommand{\arraystretch}{1.15}
\begin{tabularx}{\textwidth}{
  >{\bfseries\raggedright\arraybackslash}p{3.0cm}
  >{\raggedright\arraybackslash}p{3.3cm}
  >{\raggedright\arraybackslash}X}
\toprule
\textbf{Plane} & \textbf{Layers} & \textbf{Operational role / primary contract} \\
\midrule
Subject \& Trust &
  Identity (Ch.~6), Endpoint (Ch.~11), Crypto (Ch.~13) &
  Who is acting, from where, with what assurance. OIDC/SAML2/SCIM~2.0;
  posture schema; PKCS\#11/ACME/KMIP. Read by the Control plane. \\
\addlinespace
Execution &
  Network (Ch.~8), Compute (Ch.~9), Storage (Ch.~10) &
  Where activity runs and on which data. RADIUS + NETCONF/YANG + IPFIX;
  Kubernetes/OpenStack/OCI; S3/NFS/KMIP. Written by the Control plane. \\
\addlinespace
Control \& Governance &
  Policy (Ch.~7), Observability (Ch.~12), Orchestration (Ch.~14) &
  How decisions are made, recorded and executed. \gls{pdp} verdict
  \texttt{\{verdict, obligations, policy\_version\}}; OpenTelemetry +
  canonical TAC schema; cross-layer TAC activation (no single standard). \\
\bottomrule
\end{tabularx}
\caption{The reference architecture: nine functional layers grouped in three
planes. Two regular flows govern it --- the Subject \& Trust plane feeds the
Control plane (identity, posture and crypto parameterise policy); the Control
plane writes to the Execution plane (decisions enforced at Network, Compute,
Storage; Orchestration realises them; Observability collects evidence of every
crossing).}
\end{table}

\begin{principlebox}[Operating rule: contracts, not products]
Procurement and acceptance reference \emph{standards}, not product names. A
component is in production only when it speaks its layer's interface contract
and is \keyword{replaceable}: standard interfaces with no captive extensions
in the critical path, open state/config formats, a documented exit-data
export within an agreed window, and no hidden supplier sidecars. The one
honest gap is Orchestration: no single standard yet exists, so the
cross-layer ``glue'' is bounded deliberately to avoid creating a new captive
dependency.
\end{principlebox}

\mainref{\trchap{5}{Architectural principles} and the per-layer
chapters~6--14}

%============================================================================
\section{Migration sequence: ten waves, dependencies, reversibility}
%============================================================================

The waves are not calendar phases; they unfold under a dependency graph that
permits parallelism. \keyword{Wave~0} is the entry point for every
trajectory. After it, Waves~1--3 (Identity, Observability, Network) may run
in parallel if staffing allows; Wave~4 (Policy) needs Identity and
Observability; Wave~5 (Pilot) is the first end-to-end exercise on a contained
scope. \keyword{Every wave is reversible at acceptable cost during a published
window} --- the coexistence pattern is the reversal mechanism: as long as the
incumbent stays viable, the wave is undone by re-routing to it.

\begin{table}[h!]
\centering\small
\renewcommand{\arraystretch}{1.12}
\begin{tabularx}{\textwidth}{
  >{\bfseries\raggedright\arraybackslash}p{2.2cm}
  >{\raggedright\arraybackslash}X
  >{\raggedright\arraybackslash}p{1.5cm}
  >{\raggedright\arraybackslash}p{3.0cm}}
\toprule
\textbf{Wave} & \textbf{Operational goal} & \textbf{Requires} &
  \textbf{Reversibility window} \\
\midrule
0 --- Foundations &
  Mandate, governance body, sovereignty grid adopted, baseline inventory.
  No technical change. & --- &
  Highest; decisions informational, not binding. \\
\addlinespace
1 --- Identity &
  Sovereign IdP on OIDC/SAML2/SCIM, R\&E federation, reviewable attribute
  schema, ingestible auth log. & 0 &
  Throughout cutover; $\sim$6--12 months after final cohort. \\
\addlinespace
2 --- Observability &
  Collector layer emitting canonical TAC schema; portable ingestion;
  retention schedule. & 0 &
  Structural; remove collectors to revert. \\
\addlinespace
3 --- Network &
  Declared zone catalogue, RADIUS attachment, NETCONF/YANG config, IPFIX,
  Science~DMZ where warranted. & 0 (rec.~2) &
  Per-segment, $\sim$3--6 months after each cut-over. \\
\addlinespace
4 --- Policy &
  \gls{pdp} deployed; selected decisions migrated from app/IdP logic;
  authoring/review process. & 1, 2 (rec.~3) &
  Per-policy, while shadow observation continues. \\
\addlinespace
5 --- Pilot &
  One or more TACs in production on contained scope; extract learnings. &
  1,2,3,4 & Structural; stop the pilot at any time. \\
\addlinespace
6 --- Endpoint &
  Unified posture schema across OSes routed into the policy plane;
  decommissioning integrated. & 1, 4 (rec.~2,5) &
  Per-policy and per-population. \\
\addlinespace
7 --- Audit \& IR &
  TAC-correlated investigation, refreshed IR runbooks, routine audit
  replay. & 2,4,5 & Structurally additive; revert at boundary. \\
\addlinespace
8 --- Workload &
  Classification as authoritative input to provisioning, storage binding,
  policy; cluster-separation rule. & 1,3,4 (rec.~5) &
  Per-workload, for the coexistence duration. \\
\addlinespace
9 --- Lifecycle &
  Self-service TAC provisioning, GitOps as source of truth, stability
  metrics. & 1--8 in some form &
  Per-consumer-group. \\
\bottomrule
\end{tabularx}
\caption{Wave sequence with required pre-requisites and indicative
reversibility windows. Pace options: \emph{5-year compressed}
(parallelised, higher risk tolerance), \emph{7-year balanced} (recommended
default), \emph{10-year gradual} (renewal-aligned, lower risk). Trajectory
varies by starting estate (greenfield, brownfield single-suite-centric,
brownfield hybrid, selective, deferred).}
\end{table}

\begin{principlebox}[Minimal coherent core --- if resources are constrained]
A constrained team may legitimately adopt \keyword{less}, not later: Wave~0
(governance/inventory) + Wave~1 (sovereign identity) + Wave~2 (observability
/ audit trail), with Wave~4 (policy) the natural next step. This delivers
exactly the size-invariant, legally compelled functions (authentication,
audit, segmentation discipline under the applicable data-protection
regime --- \gls{gdpr} in the EU, with \gls{nis2} where in scope) and is a legitimate \emph{steady state}, not an abandoned
migration. Defer or mutualise the rest along the same dependency graph.
\end{principlebox}

\mainref{\trchap{17}{How to read this playbook} and the wave
chapters~18--27}

%============================================================================
\section{Conformance \& replaceability: the acceptance gates}
%============================================================================

A contract that cannot be checked is a slogan. Each wave closes by passing
the conformance subset for its scope --- never by the judgement that ``the
migration is done''. This is the single mechanism that prevents the classic
failure mode: \keyword{a product swap disguised as architectural progress}.

\begin{table}[h!]
\centering\small
\begin{tabularx}{\textwidth}{
  >{\bfseries\raggedright\arraybackslash}p{3.0cm} X}
\toprule
\textbf{Test category} & \textbf{What the gate verifies (operationally)} \\
\midrule
Interface fidelity (F) &
  The component speaks the standard correctly (e.g.\ OIDC discovery metadata
  conformant; Kubernetes upstream conformance passes; NETCONF apply without
  the vendor console). \\
\addlinespace
Data portability (P) &
  Data/config export in a documented open format and re-import into an
  alternative without loss (SCIM export to an alternative IdP; OCI image
  digest preserved; KMIP key migration). \\
\addlinespace
Exit rehearsal (X) &
  The replacement is \emph{rehearsed}, not simulated: an alternative restores
  service on a representative subset within the documented window. \\
\addlinespace
Decision auditability (A) &
  A recorded decision replays against its recorded inputs and policy version
  to the identical verdict; TAC deactivation reverses every recorded change. \\
\bottomrule
\end{tabularx}
\caption{The four acceptance-gate categories, mapped to the four conditions
of replaceability. Wave-to-test mapping is published per wave; e.g.\ Wave~1
closes against the four Identity tests, Wave~3 against the four Network
tests, Wave~5 against the union exercised by the pilot.}
\end{table}

\begin{principlebox}[How to run it]
Run a time-boxed \keyword{conformance day}: one engineer per affected layer,
one observability operator to retrieve replay material, a note-taker, the
scope owner. Half a day per layer; one day for a full review. Output is a
structured report (pass / conditional pass / fail, with evidence and
remediation owner) and an \keyword{exit-confidence index} per layer on the
0--3 grid scale. The index is a \emph{maturity trajectory}, not a day-one
gate: a layer consumed as a managed/mutualised service answers
replaceability through the provider's contractual exit terms, not a
self-run test.
\end{principlebox}

\mainref{\trchap{16}{Conformance and replaceability tests}}

%============================================================================
\section{Sizing reference points (research-intensive reference profile) for capacity planning}
%============================================================================

\begin{principlebox}[Status of these figures]
Indicative orders of magnitude for a \keyword{research-intensive}
reference profile: $\sim$25{,}000 active subjects, $\sim$1{,}500 elevated-access
staff, $\sim$500 active research projects. These are \emph{planning aids,
not measurements}; size your own estate by direct workload measurement.
\end{principlebox}

\begin{table}[h!]
\centering\small
\begin{tabularx}{\textwidth}{
  >{\bfseries\raggedright\arraybackslash}p{3.0cm}
  >{\raggedright\arraybackslash}X
  >{\raggedright\arraybackslash}p{2.6cm}}
\toprule
\textbf{Layer} & \textbf{Indicative reference points} & \textbf{Run FTE
(shared)} \\
\midrule
Identity &
  Auths/s: a few--tens steady, hundreds at peak; concurrent sessions
  thousands; \loc{identity-federation-profile} tens--hundreds of thousands/day. & 2--4 \\
\addlinespace
Policy &
  Tens--low hundreds decisions/s; $<$10\,ms p95 online decision; bundle
  refresh in minutes. & 1--2 \\
\addlinespace
Network &
  Backbone 40--400\,Gb/s aggregate; Science~DMZ at backbone order;
  IPFIX hundreds of millions of records/day. & 3--5 \\
\addlinespace
Compute / Storage &
  OpenStack + Kubernetes; Ceph multi-PB; per-class cluster separation. &
  4--7 / 2--4 \\
\addlinespace
Observability &
  Tens--hundreds of millions of events/day; tens--low hundreds of GB/day
  compressed; 90d hot / 1y warm / multi-year cold. & 3--5 \\
\bottomrule
\end{tabularx}
\caption{Selected sizing reference points and indicative run-time
engineer-equivalents per layer (ranges are \emph{not additive}; staff are
shared across layers). The build phase adds $\sim$1--2 FTE per active wave.}
\end{table}

\noindent\textbf{Scaling breakpoints} (reshape the instantiation, not the
architecture): Identity $>$\,$\sim$50k subjects $\to$ horizontal
partitioning; Audit $>$\,$\sim$500\,GB/day $\to$ hot/warm/cold tier
separation; Policy $>$\,$\sim$5{,}000 decisions/s $\to$ per-app sidecars;
Storage $>$ multi-PB $\to$ Ceph federation. \textbf{Down-scaling} reads the
same logic backward: a teaching-led institution collapses planes to logical
segments on shared hardware, mutualises compute/storage via a national
research cloud or R\&E service, and defers the optional Science~DMZ ---
a minimal-core envelope of $\sim$3--5 shared engineer-equivalents.

\mainref{\trchap{15}{Sizing reference points}}

%============================================================================
\section{Top operational risks, mitigations, and coexistence patterns}
%============================================================================

\begin{table}[h!]
\centering\small
\newcommand{\warn}{\textcolor{BrickRed}{\ding{72}}\,}%
\begin{tabularx}{\textwidth}{
  >{\raggedright\arraybackslash}p{4.4cm} X}
\toprule
\textbf{Operational risk} & \textbf{Mitigation} \\
\midrule
\warn Product swap mistaken for architectural progress &
  Close every wave on its conformance subset (F/P/X/A); a new IdP that fails
  data-portability and exit-rehearsal has not closed the wave. \\
\addlinespace
\warn Long-running dual stack with no exit date &
  Declare the exit date at wave open; track it in programme governance;
  escalate on drift. \\
\addlinespace
\warn Bridge fails silently (sync job stops, metadata expires) &
  Name the bridge owner explicitly at wave open, with a documented
  escalation path; surface discrepancies through observability. \\
\addlinespace
\warn Bidirectional sync without conflict resolution &
  Prefer unidirectional sync; where bidirectional is unavoidable, document
  and test an explicit conflict-resolution rule before production. \\
\addlinespace
\warn Coexistence frozen as de-facto architecture &
  Periodic review under the sovereignty grid; a coexistence unrevisited for
  24 months is re-evaluated against current cost/benefit. \\
\addlinespace
\warn Reversibility forfeited under delivery pressure &
  Forfeit only on explicit written acceptance by the competent governing
  authority, against the grid, revisited at every renewal. \\
\bottomrule
\end{tabularx}
\caption{Top operational risks during migration and their mitigations.}
\end{table}

\begin{principlebox}[Shadow-mode and coexistence patterns to standardise on]
\begin{itemize}[nosep,leftmargin=1.2em]
  \item \textbf{Incumbent directory $\leftrightarrow$ institutional IdP:}
    directory stays authoritative; new IdP federates via SAML2/OIDC;
    \emph{unidirectional} sync only.
  \item \textbf{Policy shadow mode:} the \gls{pdp} logs decisions with the
    same fidelity as enforcing mode before any decision becomes binding ---
    the safe path to flip enforcement.
  \item \textbf{Per-volume \gls{siem} $\leftrightarrow$ open store:} hot
    investigative window in the commercial \gls{siem}; long-tail retention in
    the open store; dashboards migrate on financial, not technical, triggers.
  \item \textbf{Commercial hypervisor $\leftrightarrow$ open substrate;
    enterprise array $\leftrightarrow$ distributed storage:} new workloads
    default to the open side; legacy migrates at its natural lifecycle
    boundary, validated by checksum.
  \item \textbf{eduroam $\leftrightarrow$ TAC-aware Wi-Fi:} keep the federated-roaming
    profile (eduroam) inviolate; add institutional/TAC attributes only on
    the non-roaming path.
\end{itemize}
\end{principlebox}

\mainref{\trchap{29}{Reversibility playbook} and
\trchap{30}{Coexistence patterns library}}

%============================================================================
\section{What this means for the operational teams}
%============================================================================

\begin{actionbox}[Operational call to action]
\begin{enumerate}[leftmargin=1.5em]
  \item \textbf{Start with Wave~0:} adopt the sovereignty grid as the
    procurement/acceptance instrument and complete the baseline inventory ---
    organisational, zero technical risk, but every later wave inherits it.
  \item \textbf{Stand up the minimal core} (Identity + Observability), then
    the Policy plane in shadow mode, before touching the network fabric.
  \item \textbf{Make conformance the definition of done:} schedule
    conformance days, publish the exit-confidence index, and refuse to close
    a wave on declaration alone.
  \item \textbf{Write procurement against contracts, not products,} and
    rehearse the exit at least once per layer per year --- replaceability is
    only real once exercised.
\end{enumerate}
\end{actionbox}

\vspace{0.4em}
\noindent\rule{\textwidth}{0.4pt}

{\footnotesize\textbf{Full document:} \emph{``Sustainable University IT --- A
Technical Reference for Architecture and Migration''}~\cite{ng-tech-ref-2026}
develops the five-dimension sovereignty grid, the Trusted Activity Context,
the nine-layer / three-plane reference architecture on open foundations, the
ten-wave migration playbook with per-wave trajectories and reversibility, the
conformance suite, the sizing reference points, and the regulatory mapping
(in the EU instantiation: \gls{gdpr}, \gls{nis2}, the EU AI Act, eIDAS~2, Schrems~II) against the
jurisdiction-neutral standards spine. It is the shared technical baseline for
the international university network anticipated in phase~2 of the SUIT
initiative.}

\smallskip

{\footnotesize\textbf{Note on the process:}
\textit{This document is an always living document. It is provided for
constructive feedback and revision in order to improve its quality. The
engineering of this document was partially supported by digital tools, some
of which incorporate Artificial Intelligence features.}}

\vspace{0.6em}
\noindent\rule{\textwidth}{0.4pt}
\printbibliography[title={\normalsize References}]

\end{document}
